### 2021-02-15 department_prez/prez.tex
progress on paper

\documentclass[dvipsnames,10pt,aspectratio=169]{beamer}
\usetheme{metropolis}           % Use metropolis theme

\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{fontspec}
\usepackage{proof}
\usepackage{tikz-cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=blue,urlcolor=blue}
\setmonofont[Scale=0.7]{DejaVu Sans Mono}

\bibliographystyle{alpha}
\setbeamerfont{bibliography item}{size=\footnotesize}
\setbeamerfont{bibliography entry author}{size=\footnotesize}
\setbeamerfont{bibliography entry title}{size=\footnotesize}
\setbeamerfont{bibliography entry location}{size=\footnotesize}
\setbeamerfont{bibliography entry note}{size=\footnotesize}
\setbeamertemplate{bibliography item}{}

%% --------------------------------------------------------------------------------

\title{Universes In Type Theory}
\date{January 29, 2021, PLC Department Workshop}
\author{\vspace{-1em}András Kovács\footnote{The author was supported by the European Union, co-financed by the
European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}
\institute{Eötvös Loránd University, Budapest}
\begin{document}
\maketitle

%% ------------------------------------------------------------

\begin{frame}{Russell's paradox, or size issues}

There is no set $S$ in a consistent set theory such that $x \in S$ iff $x \notin x$.
\vspace{1em}

Alternate formulations:
\begin{itemize}
  \item Barber's paradox
  \item No set of all sets
\end{itemize}
\vspace{1em}

Related:
\begin{itemize}
  \item Real numbers are uncountable
  \item Gödel's first incompleteness theorem
  \item Undecidability of halting
\end{itemize}
\vspace{1em}

{\footnotesize (All instances of \href{https://arxiv.org/pdf/math/0305282.pdf}{Lawvere's fixed point theorem)}}
\end{frame}

%% ------------------------------------------------------------

\begin{frame}{Solution with universes}

Solving Russell's paradox: set comprehension can only define a \textbf{subset} of a set.
\vspace{1em}

But sometimes we still want to quantify over \textbf{all sets} of
Zermelo-Fraenkel set theory.
\vspace{1em}

We can assume the existence of a very large set such that all of ZF can fit there,
which is called a \textbf{universe}. Then, just quantify over elements of the
universe. The resulting theory is stronger than ZF.
\vspace{1em}

If we want quantify over sets which \textbf{don't} fit (or universes
themselves), we can just assume even more universes.

\end{frame}

\begin{frame}[fragile]{Universes in type theory}

Agda:
\begin{verbatim}
  Bool          : Set₀
  Set₀          : Set₁
  Set₁          : Set₂
  (Bool → Set₀) : Set₁
  (Set₀ → Bool) : Set₁
\end{verbatim}

Assuming \texttt{Set₀ : Set₀} implies contradiction (by Russell-like argument).
\vspace{1em}

{\small(But type theory with \texttt{Set₀ : Set₀} is still a perfectly good Turing-complete
programming language)}
\vspace{1em}

\end{frame}

\begin{frame}[fragile]{Universes in type theory}

The basic system can be tedious:
\begin{verbatim}
  id₀ : (A : Set₀) → A → A
  id₁ : (A : Set₁) → A → A
  ...
\end{verbatim}

In Coq/Agda/Lean, various extra features are used.
Example: universe polymorphism in Agda:
\begin{verbatim}
  id : (l : Level)(A : Set l) → A → A
  id l A x = x
\end{verbatim}

\end{frame}

\begin{frame}[fragile]{Design choices and variations}

\textbf{How many universes?} Agda/Coq: countably many.
\vspace{1em}

\textbf{Are universes totally ordered?} Agda/Coq: yes.
\vspace{1em}

\textbf{What kind of level polymorphism?}. Coq: bounded polymorphism. Agda: no
bounds allowed. Bounded example:
\begin{verbatim}
  myId : (l : Level) → l < 3 → (A : Set l) → A → A
  myId l p A x = x
\end{verbatim}

\textbf{What kind of operations are available on levels?} Agda is more liberal than Coq. Example:
\begin{verbatim}
  ℕtoLevel : ℕ → Level
\end{verbatim}

\textbf{Are universes cumulative} Agda: no. Coq: yes. Cumulativity: whenever \texttt{A : Set i}, we also have \texttt{A : Set (i + 1)}.

\end{frame}

\begin{frame}{Research goals}

We want to know that each point in the design space makes sense.
\vspace{1em}

Making sense:
\begin{itemize}
  \item Logical consistency.
  \item Is the type theory a proper programming language? Programs should compute to values
        and not get randomly stuck.
  \item Is type checking decidable? (I don't plan to cover this in research).
\end{itemize}

Approach: use generic framework to cover as many features/variations as
possible. Prove that everything in the framework is sensible.

\end{frame}

\begin{frame}{Results}

Framework covers:
\begin{enumerate}
  \item Universe levels may come from any well-ordered set.
  \item Polymorphism with bounds is allowed.
  \item Universes are cumulative.
  \item Levels can be manipulated arbitrarily as program data (WIP).
\end{enumerate}

Things we need to assume in order to prove the framework sensible:
\begin{enumerate}
  \item \textbf{Two} universes.
  \item Types of certain infinitely branching trees (W-types).
\end{enumerate}

Potential further applications (in future work):
\begin{itemize}
  \item Information flow type systems (``secure'' and ``public'' levels).
  \item Staged compilation (``runtime'' and ``compile time'' levels).
\end{itemize}

\end{frame}

\begin{frame}

This talk is a side project intended for FSCD 2021 submission (in February).
\vspace{1em}

Other 2020 publications:
\begin{itemize}
  \item LICS 2020, with Ambrus: ``Large and Infinitary Quotient Inductive-Inductive Types''
  \item ICFP 2020: ``Elaboration with First-Class Implicit Function Types''
  \item TYPES 2019 post-proceedings, with Ambrus and Ambroise Lafont: ``For
    Finitary Induction-Induction, Induction is Enough''
\end{itemize}

\end{frame}

\begin{frame}

\begin{center} {\large Thank you!} \end{center}

\end{frame}

%% \begin{frame}[allowframebreaks]{References}
%%   \bibliography{references}
%% \end{frame}

\end{document}

### 2021-02-15 paper.tex
progress on paper

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Generalized Universe Hierarchies and First-Class Universe Levels}
%% \titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{András Kovács}{Eötvös Loránd University, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{}

%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
\authorrunning{A., Kovács} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases):

\Copyright{András Kovács} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}

\keywords{type theory, universes} %TODO mandatory; please add comma-separated list of keywords

%% \category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

\funding{The author was supported by the European Union,
co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

%% --------------------------------------------------------------------------------

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{amssymb}

%% \theoremstyle{definition}
%% \newtheorem{definition}{Definition}
%% \newtheorem{example}{Example}

%% \theoremstyle{theorem}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}

\theoremstyle{remark}
\newtheorem{notation}{Notation}

\theoremstyle{definition}
\newtheorem{mydefinition}{Definition}
\newtheorem{myexample}{Example}

%% Abbrevs
%% --------------------------------------------------------------------------------

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}

\renewcommand{\tt}{\mathsf{tt}}

%% --------------------------------------------------------------------------------
\begin{document}
\maketitle

\begin{abstract}
In type theories, universe hierarchies are commonly used to increase the
expressive power of the theory while avoiding inconsistencies arising from size
issues. There are numerous ways to specify universe hierarchies, and theories
may differ in details of cumulativity, choice of universe levels, specification
of type formers and eliminators, and available internal operations on levels. In
the current work, we aim to provide a framework which covers a large part of the
design space. First, we develop syntax and semantics for cumulative universe
hierarchies, where levels may come from any set equipped with a transitive
well-founded ordering. In the semantics, we show that induction-recursion can be
used to model transfinite hierarchies, and also support lifting operations on
type codes which strictly preserve type formers. Then, we consider a setup where
universe levels are first-class types and subject to arbitrary internal
reasoning. This generalizes the bounded universe polymorphism features of Coq
and also the internal level computations in Agda.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

Users of type theories often view universe levels as an unwieldy bureaucratic
detail, a necessary annoyance in service of boosting expressive power while
retaining logical consistency. However, universe hierarchies are not going away
any time soon in practical implementations of type theory. In recent
developments of systems, we are getting more universes and more adjacent
features:

\begin{itemize}
\item Agda recently added a limited version cumulativity as an optional feature
  for universes, and the upcoming 2.6.2 version will extend the $\omega+1$
  universe hierarchy to $\omega*2$.
\item Coq added support for cumulative inductive types \cite{timany18cumulative}, and added a form of
  bounded universe polymorphism \cite{ziliani15unification}.
\end{itemize}

\noindent At this point, there is a veritable zoo of universe features in existing
implementations. We have perhaps even more design choices when considering the
formal metatheory of type theories. Do type formers stay in the same
universe, or take the $\sqcup$ of universes of constituent types? Can
eliminators target any universe, or do we instead use lifting operators to cross
levels? What kind of universe polymorphism do we have, can we quantify over
bounds? Is there a type of levels, or are levels in a separate syntactic layer?

The aim of the current work is to develop semantics which covers as much as
possible from the range of sensible universe features. This way, theorists and
language implementors can grab a desired bag of features, and be able to show
consistency of their system by a straightforward translation to one of the
systems in this paper.

\paragraph*{Contributions}

\begin{enumerate}
\item In Section \ref{sec:ttgu} we describe models of type theories where
  universe levels may come from any set with a well-founded transitive ordering
  relation. We specify models as categories equipped with level-indexed diagrams
  of families, as a variation on categories with families. Each morphism of
  levels is mapped to a lifting operation on terms and types. By varying the
  preservation properties of lifting operations, we can describe a range of
  stratification features, from two-level type theory to cumulative universes.
\item We use induction-recursion to model the mentioned theories. Here, we model
  the strongest formulations for lifting and universes, namely cumulative
  universes with Russell-style type decoding. The main challenge is combining
  transfinite hierarchies with strictly type-preserving lifting operations.
\item In Section \ref{sec:ttfl} we describe type theories with internal types
  for levels and level morphisms, and extend the previous inductive-recursive
  semantics to cover these as well. Here, we can additionally represent various
  universe polymorphism features and level computations. We illustrate how
  predicative universe features in Coq and Agda could be justified in our framework.
\end{enumerate}

\noindent Several related topics are not handled in this paper.

\begin{itemize}

\item
Besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. However, we
do keep canonicity in mind while specifying our systems, and aim to avoid
pitfalls which would cause canonicity to fail.
\item
We only focus on using universes as sized-based classifiers for
types. Universe-like stratification is also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}, but we do not attempt to cover such use
cases.
\item
We only cover predicative hierarchies.
\end{itemize}

We provide an Agda formalization of the contents of the paper. The formalization
is not complete, as we skip the proofs involving an over-abundance of equality
coercions (which are more suited to informal reasoning, using equality
reflection), and instead focus on the key points.

\section{Metatheory}
\label{sec:metatheory}

We work in a Martin-Löf type which has the following features.
\begin{itemize}
  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports
    inductive-recursive types (IR) as specified by Dybjer and Setzer
    \cite{dybjer99finite}. The $\Set1$ universe is not essential and we only use
    it as a convenience feature, in this paper and in the Agda formalization. We
    may omit the universe indices if they can be inferred or if we work over
    arbitrary indices.
  \item Function extensionality and uniqueness of identity proofs
    (UIP). Additionally, we assume equality reflection in this paper, thus
    working in extensional type theory, to avoid noise from equality transports.
  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$
    inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
    $(t,\, u)$ and projections as $t.1$ and $t.2$. We have $\top$ as the unit
    type with inhabitant $\tt$, and $\bot$ as the empty type. Propositional identity
    is written as $t = u$ with $\refl : t = t$.
  \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
    function types with implicit arguments, and omit implicit applications
    whenever they can be inferred.

\end{itemize}

\section{Type Theories with Generalized Universe Hierarchies}
\label{sec:ttgu}

In this section, we first describe notions of models for type theories with
generalized universes, and discuss several variations of universes and lifting
operations. Then, we pick a concrete variant (the ``strongest'', in a sense)
and construct a model for it in the metatheory.

For the basic structure of typing contexts and substitutions, let us review
categories with families.

\begin{mydefinition}[Category with family (CwF), c.f. \cite{Dybjer96internaltype}]\label{def:cwf}
A CwF consists of the following data:
\begin{itemize}
\item A category with a terminal object. We denote the set of objects as $\Con :
  \Seti$ and use capital Greek letters starting from $\Gamma$ to refer to
  objects. The set of morphisms is $\Sub : \Con \to \Con \to \Seti$, and we use
  $\sigma$, $\delta$ and so on to refer to morphisms. The terminal object is
  $\emptycon$ with unique morphism $\epsilon : \Sub\,\Gamma\,\emptycon$. In
  initial models (that is, syntaxes) of type theories, objects correspond to
  typing contexts, morphisms to parallel substitutions and the terminal object to
  the empty context; this informs the naming scheme that we use here.
\item A \emph{family structure}, containing $\Ty : \Con \to \Seti$ and $\Tm :
  (\Gamma : \Con) \to \Ty\,\Gamma \to \Seti$, where $\Ty$ is a presheaf over the
  category of contexts and $\Tm$ is a presheaf over the category of elements of
  $\Ty$. This means that both types ($\Ty$) and terms ($\Tm$) can be
  substituted, and substitution has functorial action. We use $A$, $B$, $C$ to
  refer to types and $t$, $u$, $v$ to refer to terms, and use $A[\sigma]$ and
  $t[\sigma]$ for substituting types and terms. Additionally, a family
  structure has \emph{context comprehension} which consists of an operation
  $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$ together with an
  isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma : \Sub\,\Gamma\,\Delta)
  \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in $\Gamma$.
\end{itemize}
\end{mydefinition}

\noindent From the comprehension structure, we recover the following notions:

\begin{itemize}
\item By going right-to-left along the isomorphism, we recover \emph{substitution extension}
      $\blank,\blank : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(A[\sigma])$. This means
      that starting from $\epsilon$ or the identity substitution $\id$, we can iterate $\blank,\blank$
      to build substitutions as lists of terms.
\item By going left-to-right, and starting from $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
      the \emph{weakening substitution} $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and the \emph{zero variable}
      $\q : \Tm\,(\Gamma\ext A)\,(A[\p])$.
\item By weakening $\q$, we recover a notion of variables as De Bruijn indices. In general, the $n$-th
      De Bruijn index is defined as $\q[\p^{n}]$, where $\p^{n}$ denotes $n$-fold composition.
\end{itemize}

notions of substitution extension, weakening
and variables, internally to any CwF.

\subsection{Categories with Families}
\label{sec:categories_with_families}

\subsection{Morphisms, Lifts and Inclusions of Families}
\label{sec:morphisms}

morphisms, lifts (2ltt), inclusions, strict inclusion (cumulative)

\subsection{Level Structures}
\label{sec:level_structures}

Level structures, thin direct semicat, vs thin cat, direct semicat, direct cat

\subsection{Universes}
\label{sec:universes}

Coquand, Russell, naturality, large elimination examples
Lifting of universes
No lifting naturality for (El,Code)!

\section{Semantics}
\label{sec:semantics}

\subsection{Inductive-Recursive Codes}
\label{sec:inductive_recursive_codes}

IR codes, code lifting, properties,

\subsection{Model}

List a bunch of things in the model.

\section{Type Theories with First-Class Universe Levels}
\label{sec:ttfl}

Specification of the theory

%% --------------------------------------------------------------------------------

\bibliography{references}
\end{document}

### 2021-02-17 paper.tex
progress

@@
 \newtheorem{myexample}{Example}
+\newtheorem{mylemma}{Lemma}
 
@@
 \newcommand{\q}{\mathsf{q}}
-
+\newcommand{\Id}{\mathsf{Id}}
+\newcommand{\Nat}{\mathsf{Nat}}
+\newcommand{\Bool}{\mathsf{Bool}}
+\newcommand{\true}{\mathsf{true}}
+\newcommand{\false}{\mathsf{false}}
+\newcommand{\up}{\uparrow}
+\newcommand{\down}{\downarrow}
+\newcommand{\Lift}{\mathsf{Lift}}
 \renewcommand{\tt}{\mathsf{tt}}
+\newcommand{\Acc}{\mathsf{Acc}}
+\newcommand{\acc}{\mathsf{acc}}
+\newcommand{\Lvl}{\mathsf{Lvl}}
 
@@
 \item Agda recently added a limited version cumulativity as an optional feature
-  for universes, and the upcoming 2.6.2 version will extend the $\omega+1$
+  for universes \cite{agdadocs}, and the upcoming 2.6.2 version will extend the $\omega+1$
   universe hierarchy to $\omega*2$.
@@
   stratification features, from two-level type theory to cumulative universes.
-\item We use induction-recursion to model the mentioned theories. Here, we model
+\item We use induction-recursion to model the mentioned theories. We model
   the strongest formulations for lifting and universes, namely cumulative
@@
 We provide an Agda formalization of the contents of the paper. The formalization
-is not complete, as we skip the proofs involving an over-abundance of equality
+is not complete, as we skip proofs involving an excessive number of equality
 coercions (which are more suited to informal reasoning, using equality
@@
 
-We work in a Martin-Löf type which has the following features.
+We work in a Martin-Löf type theory which has the following features.
 \begin{itemize}
@@
     function types with implicit arguments, and omit implicit applications
-    whenever they can be inferred.
-
+    whenever they can be inferred. We may omit implicit function types
+    altogether if it is clear where certain variables are quantified.
 \end{itemize}
@@
 generalized universes, and discuss several variations of universes and lifting
-operations. Then, we pick a concrete variant (the ``strongest'', in a sense)
+operations. Then, we pick a concrete variant (the strongest, in a sense)
 and construct a model for it in the metatheory.
@@
 
-\begin{mydefinition}[Category with family (CwF), c.f. \cite{Dybjer96internaltype}]\label{def:cwf}
-A CwF consists of the following data:
+\subsection{Categories with Families}
+\label{sec:categories_with_families}
+
+\begin{mydefinition}
+A \emph{category with family} (cwf) \cite{Dybjer96internaltype} consists of the following data:
 \begin{itemize}
@@
   refer to types and $t$, $u$, $v$ to refer to terms, and use $A[\sigma]$ and
-  $t[\sigma]$ for substituting types and terms. Additionally, a family
-  structure has \emph{context comprehension} which consists of an operation
-  $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$ together with an
-  isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma : \Sub\,\Gamma\,\Delta)
-  \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in $\Gamma$.
+  $t[\sigma]$ for substituting types and terms. Additionally, a family structure
+  has \emph{context comprehension} which consists of an context extension
+  operation $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$
+  together with an isomorphism $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
+  \Sub\,\Gamma\,\Delta) \times \Tm\,\Gamma\,(A[\sigma]))$ which is natural in
+  $\Gamma$.
 \end{itemize}
@@
 
-notions of substitution extension, weakening
-and variables, internally to any CwF.
+There are other ways for presenting the basic categorical structure of models,
+which are nonetheless equivalent to cwfs, including natural models \cite{awodey18natural}
+and categories with attributes \cite{cartmellthesis}. We use the cwf presentation for its
+immediately algebraic character and closeness to syntax, as an explicit
+substitution calculus can be directly read off the specification.
 
-\subsection{Categories with Families}
-\label{sec:categories_with_families}
+\begin{notation}As De Bruijn indices are hard to read, we will mostly use
+nameful notation for binders. For example, assuming $\Nat : \{\Gamma : \Con\}
+\to \Ty\,\Gamma$ and $\Id : \{\Gamma : \Con\}(A : \Ty\,\Gamma) \to
+\Tm\,\Gamma\,A \to \Tm\,\Gamma\,A \to \Ty\,\Gamma$, we may write $\emptycon \ext
+(n : \Nat) \ext (p : \Id\,\Nat\,n\,n)$ for a typing context, instead of using
+numbered variables or cwf combinators as in $\emptycon \ext \Nat \ext
+\Id\,\Nat\,\q\,\q$.
+\end{notation}
 
-\subsection{Morphisms, Lifts and Inclusions of Families}
+\begin{notation}
+In the following, we will denote families by ($\Ty$,$\Tm$) pairs and overload context
+extension $\blank\ext\blank$ for different families.
+\end{notation}
+
+A family structure may be closed under certain \emph{type formers}. For example,
+we may close a family over function types by assuming $\Pi : (A : \Ty\,\Gamma)
+\to \Ty\,(\Gamma\ext A) \to \Ty\,\Gamma$ together with abstraction, application,
+$\beta\eta$-rules, and equations for the action of substitution on type and term
+formers.
+
+\subsection{Morphisms, Liftings and Inclusions of Families}
 \label{sec:morphisms}
 
-morphisms, lifts (2ltt), inclusions, strict inclusion (cumulative)
+In the rest of the paper we make use of categories equipped with possibly
+multiple family structures, which serves as basis for specifying universe
+hierarchies. However, it is not very useful to simply have multiple copies of
+family structures together with their type formers. In that case, every
+constructor and eliminator of every type former stays in the same family, and
+there is no interaction between families, the most we can do is to mix them
+together in typing contexts. In this subsection we describe several ways of
+crossing between families.
+
+\begin{mydefinition}
+A \emph{family morphism} $F$ between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$, $\Ty_1$)
+families consists of natural transformations mapping types to types and terms to
+terms, which preserves context extensions up to context isomorphism, i.e.\ we
+have that $(\Gamma \ext F\,A) \simeq (\Gamma \ext A)$, where $\simeq$ denotes
+existence of an invertible context morphism.
+\end{mydefinition}
+
+Family morphisms are restrictions of so-called \emph{weak morphisms}
+\cite{dependentrightadjoints} (or \emph{pseudomorphisms}
+\cite{kaposi2019gluing}) of cwfs: a weak morphism which has the identity action
+on the base category is exactly a family morphism. A plain family morphism is
+still not too interesting, so we add more features.
+
+\begin{mydefinition}
+A \emph{lifting} is a family morphism which is bijective on terms. Note that
+we can drop the $\ext$-preservation condition from the specification of
+liftings, since it follows from the invertible action on terms.
+\end{mydefinition}
+
+\begin{notation}
+We write $\Lift : \Ty_0\,\Gamma \to \Ty_1\,\Gamma$ for the action of some lifting
+on types, $\up : \Tm_0\,\Gamma\,A \to \Tm_1\,\Gamma\,(\Lift\,A)$ for the action
+on terms, and $\down$ for the inverse action on terms.
+\end{notation}
+
+\todo{Mention dependent right adjoints}
+
+\paragraph{Two-level type theory}
+Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$), and a
+lifting between them. This corresponds to a basic version of \emph{two-level
+  type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
+$\Tm_1$) as a metaprogramming layer which can generate object-level programs in
+the ($\Ty_0$, $\Tm_0$) layer.
+
+Lifted types correspond to types of object-level expressions; for
+example, $\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
+$\Lift\,\Bool_0$ is the meta-level type of $\Bool_0$-expressions, and $\Bool_1 :
+\Ty_1\,\Gamma$ is the type of meta-level Booleans. It is possible to compute a
+$\Lift\,\Bool_0$ from a $\Bool_1$. Given $b : \Tm_1\,\Gamma\,\Bool_1$, we can
+construct
+$\down(\mathsf{if}\,b\,\mathsf{then}\up\!\true_0\,\mathsf{else}\up\!\false_0) :
+\Tm_0\,\Gamma\,\Bool_0$.
+
+But there is no way to compute a $\Bool_1$ from a $\Lift\,\Bool_0$, since
+$\Lift\,\Bool_0$ has no elimination rule in $\Ty_1$. Conceptually, terms
+of $\Lift\,\Bool_0$ are \emph{expressions}, so they are not necessarily $\true$ or
+$\false$, they can also be variables or neutral expressions, so the usual
+Boolean elimination is not justified for them. In general, the setup ensures that
+we can eliminate from positive types in $\Ty_1$ to their counterparts in
+$\Ty_0$, but not the other way around, while negative types are preserved by
+$\Lift$ up to type isomorphism.
+
+Remarkably, the simple rules of two-level type theory appear to model a form of
+generative two-stage compilation with dependent types. Comparing this system to
+e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can relate $\Lift$ to $\mathsf{code}$, $\up$
+to the quasi-quotation operation $<\!\blank\!>$, and $\down$ to escape
+$\sim\!\blank$.
+
+While the staging aspect could be subject of future research, we currently focus
+on ``sizing'' hierarchies instead of staging hierarchies. This means that we
+want to eliminate from any family to any other family which is ``connected'' by
+a morphism.
+
+\begin{mydefinition} A \emph{family inclusion} is a lifting which preserves all type
+and term formers. This assumes that every type former which is contained in the source family,
+is also contained in the target family.
+\end{mydefinition}
+
+\noindent Some examples for preservation equations for type and term formers:
+\begin{alignat*}{3}
+  & \Lift\,(\Pi\,(x : A) B)   && =\,\,\,\,&& \Pi\,(x : \Lift\,A)(\Lift\, (B[x \mapsto\,\down\!x]))\\
+  & \up(\lambda\,(x : A).\,t) && =&& \lambda\,(x : \Lift\,A).\up(t[x\mapsto\down\!x])\\
+  & \Lift\,\Bool_0            && =&& \Bool_1\\
+  & \up\true_0                && =&& \true_1
+\end{alignat*}
+In general, we can skip specifying preservation for $\down$, since it follows from $\up$ preservation
+equations.
+
+Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_0$). Now, we can
+eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
+\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
+\Tm_1\,\Gamma\,(\Lift_1\,\Bool_0)$, hence $\up\!b :
+\Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
+$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
+\Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
+computes appropriately on canonical terms: if $b$ is $\true_0$, then we get
+$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.
+
+A family inclusion corresponds to a \emph{cumulative hierarchy} consisting of
+two families: every type former of the smaller family is included in the larger
+family, with the same elimination rules.
+
+\begin{mydefinition}\label{def:strict_inclusion}
+A \emph{strict family inclusion} between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$,
+$\Ty_1$) is a family inclusion for which the following equations hold:
+\begin{alignat}{2}
+  & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
+  & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
+  & \up\!t &&= t                                    \label{eq:cumlift}
+\end{alignat}
+
+%% \begin{enumerate}
+%% \item $(\Gamma \ext \Lift\,A) = (\Gamma \ext A)$
+%% \item $\Tm_1\,\Gamma\,(\Lift\,A) = \Tm_0\,\Gamma\,A$
+%% \item $\up\!t = t$
+%% \end{enumerate}
+\end{mydefinition}
+
+A strict inclusion corresponds to Sterling's \emph{algebraic cumulativity}
+\cite{sterling2019algebraic}. The additional equations are a matter of convenience: they allow us
+to omit term liftings in informal syntax\footnote{In a proof assistant, we would
+  still have to explicitly transport along the strict inclusion
+  equations.}. Most of the time we can also omit level annotations
+on term formers. For example, we have $\true_0 : \Tm_0\,\Gamma\,\Bool_0$,
+but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$, hence $\true_0 :
+\Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally equal to
+$\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using simply $\true$
+is fine whenever the family is clear from context.
+
+The definitional equality of $\true_0$ and $\true_1$ is important; without it
+canonicity would fail, since $\true_0$, $\false_0$, $\true_1$ and $\false_1$
+would be four definitionally distinct inhabitants of $\Bool_1$. See Luo
+\cite{luo2012notes} on related issues with cumulativity. This means that it is
+not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
+and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$ together
+with equation \ref{eq:cumlift} to identify term formers in different families. The
+other direction $\down\!t = t$ is immediately derivable.
 
@@
 
-Level structures, thin direct semicat, vs thin cat, direct semicat, direct cat
+We would like to describe a range of setups with multiple families and morphisms
+between them. In this subsection we describe the indexing structures for such
+family diagrams. First, some preliminary definitions.
+
+\begin{mydefinition} The \emph{accessibility predicate} on relations is defined by the following
+inductive rules:
+\begin{alignat*}{3}
+  & \Acc : \{A : \Seti\}(R : A \to A \to \Seti) \to A \to \Seti \\
+  & \acc : \{a : A\} \to ((\mi{a'} : A) \to R\,\mi{a'}\,a \to \Acc\,R\,\mi{a'}) \to \Acc\,R\,a
+\end{alignat*}
+\end{mydefinition}
+
+\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$ all
+descending $R$-chains must be finite. This is ensured by the universal property
+of the inductive definition, as all inductive types must be well-founded.
+
+\begin{mylemma} Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are propositionally equal; see
+  \cite[Section 10.3]{hottbook}.
+\end{mylemma}
+
+\begin{mydefinition} A relation $R : A \to A \to \Seti$ is \emph{well-founded} if $(a : A) \to \Acc\,R\,a$.
+\end{mydefinition}
+
+\begin{mydefinition} A \emph{level structure} consists of the following components:
+\begin{alignat*}{3}
+  & \Lvl                  &&: \Set0 \\
+  & \blank\!<\!\blank     &&: \Lvl \to \Lvl \to \Set0 \\
+  & \mathsf{<\!prop }     &&: (p\,q : i < j) \to p = q \\
+  & \blank\!\circ\!\blank &&: j < k \to i < j \to i < k \\
+  & \mathsf{<\!wf}        &&: (i : \Lvl) \to \Acc\,<\,i
+\end{alignat*}
+\end{mydefinition}
+
+\noindent We overload $\Lvl$ to refer to a given level structure and also its
+underlying set. In short, a level structure is a set together with a transitive
+well-founded relation.
+
+\begin{mydefinition}
+A \emph{family diagram} over $\Lvl$ maps each $i : \Lvl$ to a family structure
+($\Ty_i$, $\Tm_i$), and each $p : i < j$ to a strict inclusion
+($\Lift_{i}^{j}\,p$, $\up_{i}^{j}p$, $\down_{i}^{j}p$) between ($\Ty_i$,
+$\Tm_i$) and ($\Ty_j$, $\Tm_j$). Moreover, the mapping is functorial, so
+$\Lift_{i}^{k}\,(p\circ q)\,A = \Lift_{j}^{k}\,p\,(\Lift_{i}^{j}\,q\,A)$, and
+similarly for $\up_{i}^{j}p$ and $\down_{i}^{j}p$.
+\end{mydefinition}
+
+\begin{notation}
+Sometimes we omit some of the $i$, $j$, $p$ annotations from type and term
+liftings, if they are clear from context.
+\end{notation}
+
+Our choice of level structures and diagrams is motivated by the following
+reasons. First, we do not need identity morphisms in levels, because they would
+be mapped to trivial liftings, which are not interesting in our setting. Second,
+we do not need proof-relevant level morphisms, since any parallel pair of family
+liftings gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
+i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
+\Tm_i\,\Gamma\,A \simeq \Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$
+and $\Lift\,q\,A$ are in the same family, we can internally prove them
+isomorphic using function types and identity types.
+
+%% \begin{notation}
+%% Given $p : i < j$ we may write $\Lift_{i}^{j}\,p : \Ty_i\,\Gamma \to
+%% \Ty_j\,\Gamma$ for type lifting, and $\up_{i}^{j}p$ and $\down_{i}^{j}p$
+%% accordingly for term lifting. Sometimes we omit $i$, $j$ and $p$ annotations,
+%% when there is no ambiguity.
+%% \end{notation}
+
+%% \end{definition}
 

### 2021-02-17 paper.tex
progress

@@
 \newcommand{\Lvl}{\mathsf{Lvl}}
+\renewcommand{\U}{\mathsf{U}}
+\newcommand{\Code}{\mathsf{Code}}
 
@@
 
+In the following, whenever we introduce a type former, we always assume that it
+is natural with respect to substitution, i.e.\ all type and term formers have a
+corresponding substitution rule.
+
 \subsection{Morphisms, Liftings and Inclusions of Families}
@@
 constructor and eliminator of every type former stays in the same family, and
-there is no interaction between families, the most we can do is to mix them
+there is no interaction between families, and the most we can do is to mix them
 together in typing contexts. In this subsection we describe several ways of
@@
 \paragraph{Two-level type theory}
-Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$), and a
+Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
 lifting between them. This corresponds to a basic version of \emph{two-level
@@
 
-\begin{mydefinition} A \emph{family inclusion} is a lifting which preserves all type
-and term formers. This assumes that every type former which is contained in the source family,
-is also contained in the target family.
+\begin{mydefinition}\label{def:inclusion}
+A \emph{family inclusion} is a lifting which preserves all type and term
+formers. This assumes that every type former which is contained in the source
+family, is also contained in the target family.
 \end{mydefinition}
@@
 A \emph{strict family inclusion} between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$,
-$\Ty_1$) is a family inclusion for which the following equations hold:
+$\Ty_1$) is a family inclusion ($\Lift$, $\up$, $\down$) for which the following equations hold:
 \begin{alignat}{2}
@@
 \end{alignat}
-
-%% \begin{enumerate}
-%% \item $(\Gamma \ext \Lift\,A) = (\Gamma \ext A)$
-%% \item $\Tm_1\,\Gamma\,(\Lift\,A) = \Tm_0\,\Gamma\,A$
-%% \item $\up\!t = t$
-%% \end{enumerate}
 \end{mydefinition}
@@
 would be four definitionally distinct inhabitants of $\Bool_1$. See Luo
-\cite{luo2012notes} on related issues with cumulativity. This means that it is
+\cite{luo2012notes} for a discussion of related issues with cumulativity. It is
 not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
-and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$ together
-with equation \ref{eq:cumlift} to identify term formers in different families. The
-other direction $\down\!t = t$ is immediately derivable.
+and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$
+together with equation \ref{eq:cumlift} to identify term formers in different
+families. The other direction $\down\!t = t$ is immediately derivable.
 
@@
 
-\begin{mylemma} Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are propositionally equal; see
-  \cite[Section 10.3]{hottbook}.
+\begin{mylemma}
+Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
+propositionally equal \cite[Lemma 10.3.4]{hottbook}.
 \end{mylemma}
 
-\begin{mydefinition} A relation $R : A \to A \to \Seti$ is \emph{well-founded} if $(a : A) \to \Acc\,R\,a$.
+\begin{mydefinition}
+A relation $R : A \to A \to \Seti$ is \emph{well-founded} if $(a : A) \to
+\Acc\,R\,a$.
 \end{mydefinition}
+
+\begin{mylemma}
+Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
+\end{mylemma}
 
@@
 A \emph{family diagram} over $\Lvl$ maps each $i : \Lvl$ to a family structure
-($\Ty_i$, $\Tm_i$), and each $p : i < j$ to a strict inclusion
+($\Ty_i$, $\Tm_i$), and each $p : i < j$ to a family inclusion
 ($\Lift_{i}^{j}\,p$, $\up_{i}^{j}p$, $\down_{i}^{j}p$) between ($\Ty_i$,
@@
 $\Lift_{i}^{k}\,(p\circ q)\,A = \Lift_{j}^{k}\,p\,(\Lift_{i}^{j}\,q\,A)$, and
-similarly for $\up_{i}^{j}p$ and $\down_{i}^{j}p$.
+similarly for $\up_{i}^{j}p$ and $\down_{i}^{j}p$. A \emph{strict family diagram}
+is a family diagram where each inclusion is strict.
 \end{mydefinition}
@@
 and $\Lift\,q\,A$ are in the same family, we can internally prove them
-isomorphic using function types and identity types.
-
-%% \begin{notation}
-%% Given $p : i < j$ we may write $\Lift_{i}^{j}\,p : \Ty_i\,\Gamma \to
-%% \Ty_j\,\Gamma$ for type lifting, and $\up_{i}^{j}p$ and $\down_{i}^{j}p$
-%% accordingly for term lifting. Sometimes we omit $i$, $j$ and $p$ annotations,
-%% when there is no ambiguity.
-%% \end{notation}
-
-%% \end{definition}
+isomorphic using function types and identity types. That said, every construction
+in this paper would still work with inverse categories as level structures.
 
@@
 
-Coquand, Russell, naturality, large elimination examples
-Lifting of universes
-No lifting naturality for (El,Code)!
+At this point, we can talk about family diagrams, but no previously seen type
+former depends on levels in an interesting way. For example, $\Bool_i$ has the
+same canonical inhabitants as $\Bool_j$, for any $i$ and $j$. Universes
+introduce dependency on levels, by serving as classifiers for smaller families internally
+to larger families.
+
+\todo{find reference for Coquand univ}
+
+\begin{mydefinition} A family diagram supports \emph{universe formation} if it supports the following:
+\begin{alignat*}{3}
+  & \U             &&: (i\,j : \Lvl) \to i < j \to \Ty_j\,\Gamma\\
+  & \mathsf{LiftU} &&: \Lift_{j}^{k}\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)
+\end{alignat*}
+\end{mydefinition}
+We also need a way to pin down universes as classifiers. We consider two variants.
+\begin{mydefinition}
+A family diagram has \emph{Coquand universes} if it has universe formation and
+additionally supports $\El : \Tm_j\,\Gamma\,(\U\,i\,j\,p) \to \Ty_i\,\Gamma$, and
+its inverse $\Code : \Ty_i\,\Gamma \to \Tm_j\,\Gamma\,(\U\,i\,j\,p)$.
+\end{mydefinition}
+\begin{mydefinition}
+A family diagram has \emph{Russell universes} if it has Coquand universes and
+additionally satisfies $\Tm_j\,\Gamma\,(\U\,i\,j\,p) = \Ty_i\,\Gamma$ and
+$\El\,t = t$.
+\end{mydefinition}
+
+The move from Coquand to Russell universes is fairly similar to the move from
+inclusions to strict inclusions. The Russell variant makes it possible to
+informally omit $\El$ and $\Code$. Likewise, the $\El\,t = t$ condition ensures
+appropriate naturality. If we only assumed $\Tm_j\,\Gamma\,(\U\,i\,j\,p) =
+\Ty_i\,\Gamma$ but not Coquand universes, we would not be able to prove that a
+$t : \Tm_j\,\Gamma\,(\U\,i\,j\,p)$ substituted \emph{as a term} is the same
+thing as $t$ substituted \emph{as a type}. Both would be written as $t[\sigma]$
+in our notation, but they involve different $\blank[\blank]$ operations.
+
+Unlike every other type or term former, there is no lifting computation rule for
+$\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
+lifting and term lifting, but while term lifting has invertible action, type
+lifting does not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a
+larger $\Ty$ is not isomorphic a smaller one, because it contains more
+universes. So, for example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to
+$\Ty_1\,\Gamma$ yields $\Bool_1$, but lifting $\Bool_0$ as a term yields
+$\Bool_0$.
+
+Assuming Coquand or Russell universes and $p : i < j$, we can recover
+polymorphic functions, for example, we may have $\mi{id} : \Pi(A : \U\,i\,j\,p)
+(\Lift\,p\,(\El\,A) \to \Lift\,p\,(\El\,A))$ for the polymorphic identity. Here,
+we quantify over terms of $\U$, and since every type former stays on the same
+level (including $\Pi$), we have to $\Lift$ the types in the codomain to the
+level of the domain. We can also recover large elimination, for example as in
+\[
+(\lambda\,(b : \Bool_j).\,\mathsf{if}\,b\,\mathsf{then}\,\Code\,\top_i\,\mathsf{else}\,\Code\,\bot_i)
+: \Tm_j\,\Gamma\,(\Bool_j \to \U\,i\,j\,p).
+\]
 
@@
 \label{sec:semantics}
+
+In this section we give a model for a type theory with generalized
+universes. Let's make the notion of model concrete first.
+
+\begin{mydefinition} Fix a $\Lvl$ structure. A model for a type theory with $\Lvl$-indexed
+universes consists of the following:
+\begin{enumerate}
+  \item A base category ($\Con$, $\Sub$) with a terminal object $\emptycon$.
+  \item A strict family diagram ($\Ty_i$, $\Tm_i$) over $\Lvl$, supporting Russell
+        universes, and each family structure is closed under the same basic type formers.
+\end{enumerate}
+\end{mydefinition}
+
+We abbreviate the theory as TTGU, for ``type theory with generalized
+universes''. The choice of available basic type formers is up to personal taste,
+and it will not significantly affect the following model construction.
 

### 2021-02-18 paper.tex
progress

@@
 \newcommand{\Code}{\mathsf{Code}}
+\newcommand{\msf}[1]{\mathsf{#1}}
+\newcommand{\uir}{\msf{U^{IR}}}
+\newcommand{\elir}{\msf{El^{IR}}}
+\newcommand{\ult}{\U_{<}}
 
@@
   the strongest formulations for lifting and universes, namely cumulative
-  universes with Russell-style type decoding. The main challenge is combining
-  transfinite hierarchies with strictly type-preserving lifting operations.
+  universes with Russell-style type decoding.
 \item In Section \ref{sec:ttfl} we describe type theories with internal types
@@
   universe polymorphism features and level computations. We illustrate how
-  predicative universe features in Coq and Agda could be justified in our framework.
+  predicative universe features in Coq and Agda can be justified in our framework.
 \end{enumerate}
@@
 
-\section{Type Theories with Generalized Universe Hierarchies}
+\section{Generalized Universe Hierarchies}
 \label{sec:ttgu}
@@
 
-\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$ all
+\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$, all
 descending $R$-chains must be finite. This is ensured by the universal property
@@
 
-\begin{mylemma}
+\begin{mylemma}\label{lem:accprop}
 Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
-propositionally equal \cite[Lemma 10.3.4]{hottbook}.
+equal \cite[Lemma 10.3.4]{hottbook}.
 \end{mylemma}
@@
 $\El$ and $\Code$. Intuitively, the issue is that we would need to relate type
-lifting and term lifting, but while term lifting has invertible action, type
-lifting does not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a
-larger $\Ty$ is not isomorphic a smaller one, because it contains more
-universes. So, for example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to
-$\Ty_1\,\Gamma$ yields $\Bool_1$, but lifting $\Bool_0$ as a term yields
-$\Bool_0$.
+lifting and term lifting, but while term lifting is invertible, type lifting is
+not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a larger $\Ty$ is
+not isomorphic a smaller one, because it contains more universes. So, for
+example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to $\Ty_1\,\Gamma$ yields
+$\Bool_1$, but lifting $\Bool_0$ as a term yields $\Bool_0$.
 
@@
 polymorphic functions, for example, we may have $\mi{id} : \Pi(A : \U\,i\,j\,p)
-(\Lift\,p\,(\El\,A) \to \Lift\,p\,(\El\,A))$ for the polymorphic identity. Here,
-we quantify over terms of $\U$, and since every type former stays on the same
-level (including $\Pi$), we have to $\Lift$ the types in the codomain to the
-level of the domain. We can also recover large elimination, for example as in
+(\Lift\,p\,(\El\,A) \to \Lift\,p\,(\El\,A))$ for the polymorphic identity
+function. Here, we quantify over terms of $\U$, and since every type former
+stays on the same level (including $\Pi$), we have to $\Lift$ the types in the
+codomain to match the level of the domain. We can also recover large
+elimination, for example as in
 \[
@@
 \]
+
+\todo{ordinal indexed levels and lubs}
 
@@
 
-\begin{mydefinition} Fix a $\Lvl$ structure. A model for a type theory with $\Lvl$-indexed
-universes consists of the following:
+\begin{mydefinition}[Notion of model for a type theory with generalized universes (TTGU)]
+Fix a $\Lvl$ structure. A model for TTGU consists of
 \begin{enumerate}
@@
 
-We abbreviate the theory as TTGU, for ``type theory with generalized
-universes''. The choice of available basic type formers is up to personal taste,
-and it will not significantly affect the following model construction.
+The choice of available basic type formers is up to personal taste, and it will
+not significantly affect the following model construction.
+
+Both in families and universes we choose the stricter formulation, since if we
+give a model which proves the strict syntax consistent, we immediately get a model which
+proves the weak syntax consistent\footnote{We always get \emph{initial} and \emph{terminal}
+models automatically, because of the algebraic character of the theories in this paper. We
+also get a \emph{freely generated} strict model from a weak model, from the left adjoint
+of the functor which forgets the strictness equations. But none of these tricks can be used
+to automatically get a consistency proof.}.
 
@@
 
-IR codes, code lifting, properties,
+The task is to interpret the $\Lvl$-many universes of TTGU using a metatheoretic
+feature which is more compact and already well-studied. For this, we need to
+define a $\Lvl$-indexed type of type codes. Since $\Lvl$ and $\!\blank<\!\blank$
+can be arbitrary, we effectively need to define transfinite hierarchies of
+codes. We use an inductive-recursive (IR) \cite{dybjer99finite} definition, for
+the following reasons.
 
-\subsection{Model}
+First, IR is already supported in the proof assistant Agda, and it is very
+useful to be able to sketch out ideas in a machine-checked setting. It would be
+much harder to do the same when developing semantics in set theory.
 
-List a bunch of things in the model.
+Second, could we use weaker type-theoretic features, like super universes
+\cite{Palmgren98onuniverses} or Mahlo universes \cite{setzer00mahlo}? These are
+sufficient to model transfinite hierarchies. However, it is not clear how to
+additionally support \emph{recursive sub-universes}, which in our work corresponds to
+the type former preservation property of $\Lift$. Thus, we give a custom IR
+definition which corresponds more directly to TTGU structure.
 
-\section{Type Theories with First-Class Universe Levels}
+\begin{mydefinition}[IR codes]
+  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define
+  $\uir$ and $\elir$ by simultaneous IR. For basic type formers, below we only
+  include codes for function types and Booleans; other type formers are
+  straightforward to add (and we do have more in our Agda formalization).
+\begin{alignat*}{5}
+  & \uir   &&: \Set0
+     && \rlap{$\elir : \uir \to \Set0$} \\
+  & \U'    &&: (j : \Lvl) \to j < i \to \uir
+     && \elir\,(\U'\,j\,p) &&= f\,j\,p \\
+  & \Pi'   &&: (A : \uir) \to (\elir\,A \to \uir) \to \uir \hspace{2em}
+     && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
+  & \Bool' &&: \uir
+     && \elir\,\Bool'      &&= \Bool
+\end{alignat*}
+\end{mydefinition}
+
+\begin{notation}
+We may write $\uir_{i\,f}$ and $\elir_{i\,f}$ in order to make parameters explicit.
+\end{notation}
+
+($\uir$, $\elir$) can be viewed as a \emph{universe operator}: given semantics
+for an initial segment of $\Lvl$ (given by $i$ and $f$), we create a new
+universe which is closed under basic type formers, and also closed under all
+sets in $f$ by the way of $\U'$. Most importantly, this operation can be
+transfinitely iterated. We first define universes for initial segments of
+$\Lvl$, by induction on the accessibility of levels:
+\begin{alignat*}{3}
+  & \ult\,: \{i : \Lvl\}\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
+  & \ult\,\{j\}\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})}
+\end{alignat*}
+
+\begin{mydefinition}[Semantic universe]
+Since every level is accessible, we can define the full semantic hierarchy and
+its decoding function.
+\begin{alignat*}{3}
+  &\U : \Lvl \to \Set0
+     && \El : \{i : \Lvl\} \to \U\,i \to \Set0 \\
+  &\U\,i = \uir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}\hspace{1em}
+     && \El\,\{i\} = \elir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}
+\end{alignat*}
+\end{mydefinition}
+
+\begin{mylemma}\label{lem:ucomp}
+Assuming $p : i < j$, we have the computation rule $\ult\,j\,p = \U\,j$. Proof: we
+may assume that any witness for $\Acc\,<\,i$ is of the form $\acc\,f$ for
+some $f$. Then the equation becomes $\uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})} =
+\uir_{j\,(\ult\,\{j\}\,\{\mathsf{<\!wf}\,j\})}$, but by Lemma \ref{lem:accprop}
+the $f\,j\,p$ and $\mathsf{<\!wf}\,j$ witnesses are equal. \qed
+\end{mylemma}
+
+\begin{mydefinition}[Semantic $\Lift$]
+We define by induction on $\uir$ a function with type $(p : i < j)(A : \U\,i)
+\to (A' : \U\,j)\times(\El\,A' = \El\,A)$. However, for the sake of clarity, we
+present this here as two (mutual) functions:
+\begin{alignat*}{3}
+  &\Lift\,      &&: (p : i < j) \to \U\,i \to \U\,j\\
+  &\msf{ElLift} &&: (p : i < j)(A : \U\,i) \to \El\,(\Lift\,A) = \El\,A
+\end{alignat*}
+Let us look at $\Lift$ first:
+\begin{alignat*}{3}
+  &\Lift\,p\,(\U'\,k\,q)    &&= \U'\,k\,(p \circ q)\\
+  &\Lift\,p\,(\Pi'\,A\,B)   &&= \Pi'\,(\Lift\,p\,A)\,(\lambda\,a.\,\Lift\,p\,(B\,a))\\
+  &\Lift\,p\,\Bool'         &&= \Bool'
+\end{alignat*}
+Above, the $\Pi'$ definition is well-typed by $\msf{ElLift}\,p\,A$. For the
+proof of $\msf{ElLift}$, the only interesting case is $\U'$. Here, we need
+show $\ult\,\{j\}\,k\,(p \circ q) = \ult\,\{i\}\,k\,q$, but by Lemma
+\ref{lem:ucomp} both sides are $\U\,k$. \qed
+\end{mydefinition}
+
+\begin{mylemma} Properties of $\Lift$:
+\begin{enumerate}
+  \item $\Lift$ preserves all basic type formers; this is immediate from the definition.
+  \item $\Lift$ is functorial, i.e.\ $\Lift\,(p \circ q)\,A = \Lift\,p\,(\Lift\,q\,A)$. This follows
+    by induction on $A$, and we make use of the irrelevance of $\blank\!<\!\blank$ in the $\U'$ case.
+    \qed
+\end{enumerate}
+\end{mylemma}
+
+\subsection{Model of TTGU}
+
+We give a model of TTGU in this section.
+
+\todo{disambiguate IR and TTGU type notation}
+
+\begin{mydefinition}[Base category]
+This is simply the category of sets and functions in $\Set0$, i.e.\ $\Con = \Set0$,
+$\Sub\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
+\end{mydefinition}
+
+\begin{mydefinition}[Family diagram]
+We map $i : \Lvl$ to a family structure as follows.
+\begin{alignat*}{5}
+  &\Ty_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \Tm_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
+\end{alignat*}
+Type and term substitution are given by composition with some $\sigma : \Gamma
+\to \Delta$ function. Context extension is given by $\Gamma \ext A = (\gamma :
+\Gamma) \times \El\,(A\,\gamma)$. Since we define $\blank\!\ext\!\blank$ as a
+$\Sigma$-type, the comprehension structure for it follows immediately, as
+$\Sigma$ pairing and projections.
+
+Mapping $i$ to a strict inclusion is defined as follows. We
+\end{mydefinition}
+
+\section{First-Class Universe Levels}
 \label{sec:ttfl}

### 2021-02-19 paper.tex
progress

@@
 \usepackage{amssymb}
+\usepackage{bm}
 
@@
 
-\begin{mylemma} Properties of $\Lift$:
+\begin{mylemma}\label{lem:liftprop} Properties of $\Lift$:
 \begin{enumerate}
@@
 
-\todo{disambiguate IR and TTGU type notation}
+\begin{notation} To avoid name clashing between components of the model and metatheoretic
+definitions, we use \textbf{bold} font to refer to TTGU components.
+\end{notation}
 
 \begin{mydefinition}[Base category]
-This is simply the category of sets and functions in $\Set0$, i.e.\ $\Con = \Set0$,
-$\Sub\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
+This is simply the category of sets and functions in $\Set0$, i.e.\ $\bm{\Con} = \Set0$,
+$\bm{\Sub}\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
 \end{mydefinition}
@@
 \begin{alignat*}{5}
-  &\Ty_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \Tm_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
+  &\bm{\Ty}_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \bm{\Tm}_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
 \end{alignat*}
 Type and term substitution are given by composition with some $\sigma : \Gamma
-\to \Delta$ function. Context extension is given by $\Gamma \ext A = (\gamma :
-\Gamma) \times \El\,(A\,\gamma)$. Since we define $\blank\!\ext\!\blank$ as a
-$\Sigma$-type, the comprehension structure for it follows immediately, as
-$\Sigma$ pairing and projections.
+\to \Delta$ function. Comprehension structure is given by $\Gamma \bm{\ext} A =
+(\gamma : \Gamma) \times \El\,(A\,\gamma)$. Type lifting along $p : i < j$ is
+as follows:n
+\begin{alignat*}{5}
+  & \bm{\Lift}_{i}^{j}p : \bm{\Ty}_i\,\Gamma \to \bm{\Ty}_j\,\Gamma\\
+  & \bm{\Lift}_{i}^{j}p\,A = \lambda\,\gamma.\,\Lift_{i}^{j}p\,(A\,\gamma)
+\end{alignat*}
+Now, two of the strict inclusion equations follow from $\msf{ElLift}$, namely
+$(\Gamma \bm{\ext} \bm{\Lift}_{i}^{j}p\,A) = (\Gamma \bm{\ext} A)$ and
+$\bm{\Tm}_j\,\Gamma\,(\bm{\Lift}_{i}^{j}p\,A) = \bm{\Tm}_i\,\Gamma\,A$. Thus, we
+can just define term lifting as $\bm{\up}_{i}^{j}\!p\,t = t$ and
+$\bm{\down}_{i}^{j}\!p\,t = t$. Basic type formers are as follows.
+\begin{alignat*}{5}
+  & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\Gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
+  & \hspace{2em}\bm{\Bool} = \lambda\,\gamma.\,\Bool'
+\end{alignat*}
+$\bm{\Lift}_{i}^{j}p$ preserves type formers by Lemma \ref{lem:liftprop}. We
+define basic term formers using metatheoretic term formers, e.g.\ $\bm{\true}_i
+= \lambda\,\gamma.\,\true$ and $(\bm{\lambda}_i\,x.\,t) =
+\lambda\,\gamma\,\alpha.\,t\,(\gamma,\,\alpha)$. Note that since semantic term
+formers are just external constructors, they do not depend on levels, so
+e.g.\ $\bm{\true}_i$ is the same at all $i$. This implies that
+$\bm{\up}_{i}^{j}\!p$ preserves term formers as well, so ($\bm{\Lift}_{i}^{j}p$,
+$\bm{\up}_{i}^{j}\!p$, $\bm{\down}_{i}^{j}\!p$) is a strict family inclusion.
 
-Mapping $i$ to a strict inclusion is defined as follows. We
+We define universes as $\bm{\U}\,i\,j\,p = \lambda\,\gamma.\,\U'_i\,j\,p$.  With
+this, $\bm{\Lift}_{j}^{k}\,p\,(\bm{\U}\,i\,j\,q) = \bm{\U}\,i\,k\,(p \circ q)$
+follows by the definition of semantic $\Lift$. The Russell universe equation
+$\bm{\Tm}_j\,\Gamma\,(\bm{\U}\,i\,j\,p) = \bm{\Ty}_i\,\Gamma$ follows from Lemma
+\ref{lem:ucomp}, so we can define $\bm{\El}$ and $\bm{\Code}$ as identity functions.
+
+\begin{theorem}[Consistency of TTGU]
+There is no closed syntactic term of $\bot_i$ for any $i$.
+\end{theorem}
+\begin{proof}
+Assuming a syntactic $t : \Tm_i\,\emptycon\,\bot_i$, we can interpret it in the
+previously given model, which yields an inhabitant of the metatheoretic $\bot$,
+hence a contradiction.
+\end{proof}
+
+%% To verify that we have strict inclusions, we need to
+%% show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
+%% formers, but first we look at the semantics of type and term formers.
+
+%% To verify that we have strict inclusions, it
+%% remains to show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
+%% formers.  For $\bm{\Lift}$ this follows from Lemma \ref{lem:liftprop}. For
+%% $\bm{\up}$, note that semantics terms
+
+  %% & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
+  %% & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
+  %% & \up\!t &&= t                                    \label{eq:cumlift}
+
+%% The strict inclusions are given as follows.
+  %% Lift : ∀ {Γ i j} → i < j → Ty Γ i → Ty Γ j
+  %% Lift p A γ = ↑ p (A γ)
+
 \end{mydefinition}

### 2021-02-20 paper.tex
progress

@@
 \newcommand{\ult}{\U_{<}}
+\newcommand{\mkMor}{\msf{mk}\!_<}
+\newcommand{\unMor}{\msf{un}\!_<}
+\newcommand{\mkLvl}{\msf{mk}_{\Lvl}}
+\newcommand{\unLvl}{\msf{un}_{\Lvl}}
 
@@
   semantics to cover these as well. Here, we can additionally represent various
-  universe polymorphism features and level computations. We illustrate how
-  predicative universe features in Coq and Agda can be justified in our framework.
+  universe polymorphism features and level computations.
 \end{enumerate}
@@
 We only cover predicative hierarchies.
+\item
+We do not discuss ergonomics and implementation of elaboration. Our goal is to
+develop a system which is as flexible as possible, even though that may not be
+the best suited for inference and elaboration. For example, first-class levels
+have undecidable ordering in general.
 \end{itemize}
@@
 \label{sec:metatheory}
+
+\todo{purge unused features}
 
@@
 
-\todo{ordinal indexed levels and lubs}
-
 \section{Semantics}
@@
 \begin{mydefinition}[IR codes]
-  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define
-  $\uir$ and $\elir$ by simultaneous IR. For basic type formers, below we only
-  include codes for function types and Booleans; other type formers are
-  straightforward to add (and we do have more in our Agda formalization).
+  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$
+  and $\elir$ by simultaneous IR. For basic type formers, below we only include
+  codes for function types, the empty type, and $\Bool$; other type formers are
+  straightforward to add (and we do have more in the Agda formalization).
 \begin{alignat*}{5}
@@
      && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
+  & \bot' &&: \uir
+     && \elir\,\bot'      &&= \bot\\
   & \Bool' &&: \uir
@@
 \begin{alignat*}{3}
-  & \ult\,: \{i : \Lvl\}\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
-  & \ult\,\{j\}\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})}
+  & \ult\,: (i : \Lvl)\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
+  & \ult\,i\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,j\,\{f\,j\,p\})}
 \end{alignat*}
@@
      && \El : \{i : \Lvl\} \to \U\,i \to \Set0 \\
-  &\U\,i = \uir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}\hspace{1em}
-     && \El\,\{i\} = \elir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}
+  &\U\,i = \uir_{i\,(\ult\,i\,\{\mathsf{<\!wf}\,i\})}\hspace{1em}
+     && \El\,\{i\} = \elir_{i\,(\ult\,i\,\{\mathsf{<\!wf}\,i\})}
 \end{alignat*}
@@
 may assume that any witness for $\Acc\,<\,i$ is of the form $\acc\,f$ for
-some $f$. Then the equation becomes $\uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})} =
-\uir_{j\,(\ult\,\{j\}\,\{\mathsf{<\!wf}\,j\})}$, but by Lemma \ref{lem:accprop}
+some $f$. Then the equation becomes $\uir_{j\,(\ult\,j\,\{f\,j\,p\})} =
+\uir_{j\,(\ult\,j\,\{\mathsf{<\!wf}\,j\})}$, but by Lemma \ref{lem:accprop}
 the $f\,j\,p$ and $\mathsf{<\!wf}\,j$ witnesses are equal. \qed
@@
   &\Lift\,p\,(\Pi'\,A\,B)   &&= \Pi'\,(\Lift\,p\,A)\,(\lambda\,a.\,\Lift\,p\,(B\,a))\\
+  &\Lift\,p\,\bot'          &&= \bot'\\
   &\Lift\,p\,\Bool'         &&= \Bool'
@@
 proof of $\msf{ElLift}$, the only interesting case is $\U'$. Here, we need
-show $\ult\,\{j\}\,k\,(p \circ q) = \ult\,\{i\}\,k\,q$, but by Lemma
+show $\ult\,j\,k\,(p \circ q) = \ult\,i\,k\,q$, but by Lemma
 \ref{lem:ucomp} both sides are $\U\,k$. \qed
@@
 
-\subsection{Model of TTGU}
+\subsection{IR Model of TTGU}
 
@@
 (\gamma : \Gamma) \times \El\,(A\,\gamma)$. Type lifting along $p : i < j$ is
-as follows:n
+as follows:
 \begin{alignat*}{5}
@@
   & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\Gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
-  & \hspace{2em}\bm{\Bool} = \lambda\,\gamma.\,\Bool'
+  & \hspace{2em}\bm{\bot}_i = \lambda\,\gamma.\,\bot'
+  & \hspace{2em}\bm{\Bool}_i = \lambda\,\gamma.\,\Bool'
 \end{alignat*}
 $\bm{\Lift}_{i}^{j}p$ preserves type formers by Lemma \ref{lem:liftprop}. We
-define basic term formers using metatheoretic term formers, e.g.\ $\bm{\true}_i
+define basic term formers and eliminators using metatheoretic features, e.g.\ $\bm{\true}_i
 = \lambda\,\gamma.\,\true$ and $(\bm{\lambda}_i\,x.\,t) =
@@
 
+\end{mydefinition}
+
 \begin{theorem}[Consistency of TTGU]
@@
 
-%% To verify that we have strict inclusions, we need to
-%% show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
-%% formers, but first we look at the semantics of type and term formers.
-
-%% To verify that we have strict inclusions, it
-%% remains to show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
-%% formers.  For $\bm{\Lift}$ this follows from Lemma \ref{lem:liftprop}. For
-%% $\bm{\up}$, note that semantics terms
-
-  %% & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
-  %% & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
-  %% & \up\!t &&= t                                    \label{eq:cumlift}
-
-%% The strict inclusions are given as follows.
-  %% Lift : ∀ {Γ i j} → i < j → Ty Γ i → Ty Γ j
-  %% Lift p A γ = ↑ p (A γ)
-
-\end{mydefinition}
-
 \section{First-Class Universe Levels}
@@
 
-Specification of the theory
+In the following, we specify and model type theories where levels and
+their morphisms are represented by internal types.
+
+However, it would be awkward to pick a particular structure for levels, and
+specify a type theory which internalizes that structure; for example
+internalizing levels as natural numbers. We do not want to repeat the
+specification and semantics for each choice of level structure; instead, we aim
+to have a more generic solution.
+\begin{enumerate}
+  \item We first give a specification of \emph{type theory with dependent
+        levels}, or TTDL, where levels and level morphism may depend on typing contexts.
+        Here, liftings, universes and type formers are specified, but the internal
+        structure of levels is not yet pinned down.
+  \item We show that we can extend TTDL with \emph{level reflection} rules, which identify
+        levels with particular internal types, thereby getting \emph{type theories with first-class
+        levels}, or TTFL.
+\end{enumerate}
+This decreases the amount of work that we have to do, in order to get semantics
+for different level setups. We only need to pick an external level structure
+such that it can be also represented using TTDL type formers.
+
+\begin{mydefinition} A model of TTDL consists of the following.
+\begin{enumerate}
+\item
+  A base category ($\Con$, $\Sub$) with terminal object $\emptycon$.
+\item
+  A ``dependent'' level structure on the base category:
+  \begin{alignat*}{3}
+    &\Lvl                  &&: \Con \to \Seti\\
+    &\blank\!<\!\blank     &&: \{\Gamma : \Con\} \to \Lvl\,\Gamma \to \Lvl\,\Gamma \to \Seti\\
+    &<\!\msf{prop}         &&: (p\,q : i < j) \to p = q\\
+    &\blank\!\circ\!\blank &&: j < k \to i < j \to i < k
+  \end{alignat*}
+  Additionally, $\Lvl$ and $\blank\!<\!\blank$ are natural in the base category,
+  so they support substitution operations. \emph{Remark:} at this point, we do not require well-foundedness
+  for $\blank\!<\!\blank$, as it has no bearing on basic lifting and universe rules, and
+  well-foundedness will be usually internally provable when we add level reflection rules.
+\item
+  A ``bootstrapping'' assumption on levels. This can be any collection of levels
+  and morphisms. We pick the assumption that $l_0, l_1 : \Lvl\,\Gamma$ exist
+  together with $l_{01} : l_0 < l_1$. We need to assume at least some levels,
+  because otherwise the TTDL syntax is trivial; it is not even possible to
+  write down a non-empty context or a closed type former. $l_0 < l_1$ allows large
+  eliminations on type formers so it provides a fair amount of expressive power.
+\item
+  A family structure:
+  \begin{alignat*}{3}
+    &\Ty &&: (\Gamma : \Con) \to \Lvl\,\Gamma \to \Seti\\
+    &\Tm &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Seti\\
+    &\blank\ext\blank &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Con
+  \end{alignat*}
+  We have type and term substitution, which depends on level substitution. For instance, we have
+  \[
+    \blank[\blank] : \Ty\,\Delta\,i \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Ty\,\Gamma\,(i[\sigma])
+  \]
+  We also have a comprehension isomorphism $\Sub\,\Gamma\,(\Delta \ext A) \simeq
+  (\sigma : \Sub\,\Gamma\,\Delta)\times \Tm\,\Gamma\,(A[\sigma])$, which is
+  natural in $\Gamma$.
+\item A lifting structure with
+  \begin{alignat*}{3}
+    &\Lift &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\} \to i < j \to \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j\\
+    &\up   &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\}(p : i < j) \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,(\Lift\,p\,A)
+  \end{alignat*}
+  Such that
+  \vspace{1em}
+  \begin{enumerate}
+    \item $\Lift$ preserves all basic type formers and has functorial action on $p \circ q$.
+    \item $\up$ has an inverse $\down$, preserves all basic term formers and has functorial action on $p \circ q$.
+    \item $(\Gamma \ext A) = (\Gamma \ext \Lift\,p\,A)$, and $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$ and $\up\,t = t$.
+    %% \item $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$.
+    %% \item $\up\,t = t$.
+  \end{enumerate}
+  \vspace{1em}
+  Above we mention basic type formers, although we haven't yet specified
+  those. The way this should be understood, is that any basic type former introduced
+  from now on should come equipped with preservation equations for lifting. This is
+  similar to how we mandate that any introduced type former must be natural with respect
+  to substitution.
+\item A universe structure
+  \begin{alignat*}{6}
+    &\U  &&: \{\Gamma : \Con\}(i\,j : \Lvl\,\Gamma) \to i < j \to \Ty\,\Gamma\,j\hspace{2em}
+    &\El &&: \Tm\,\Gamma\,(\U\,i\,j\,p) \to \Ty\,\Gamma\,i
+  \end{alignat*}
+  Such that $\Lift\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)$, $\El$ has
+  inverse $\Code$, $\Tm\,\Gamma\,(\U\,i\,j\,p) = \Ty\,\Gamma\,i$ and $\El\,t =
+  t$.
+\item Basic type formers.
+\end{enumerate}
+\end{mydefinition}
+
+\begin{mydefinition}[IR model of TTDL]\label{def:ttdlmodel}
+ Assume an external $\Lvl$ structure such that it supports $l_0, l_1 : \Lvl$ and
+ $l_{01} : l_0 < l_1$ (the bootstrapping assumption). We again use the IR
+ universe constructions from Section \ref{sec:inductive_recursive_codes},
+ instantiated to the assumed $\Lvl$ structure. We describe components of the
+ model in order. Again, we write components of the model in \textbf{bold} font.
+\begin{enumerate}
+\item The base category remains unchanged from the TTGU model.
+\item For the level structure, we define $\bm{\Lvl}\,\Gamma = \Gamma \to \Lvl$ and
+  $i \bm{<} j = (\gamma : \Gamma) \to i\,\gamma < j\,\gamma$. Subsitution for internal
+  levels and morphisms is given by function composition with $\sigma : \Gamma \to \Delta$.
+  Internal composition and $\bm{<\!\msf{prop}}$ follow from the external counterparts.
+\item The internal bootstrapping assumption is modeled with the external counterpart.
+\item We define $\bm{\Ty}\,\Gamma\,i = (\gamma : \Gamma) \to \U\,(i\,\gamma)$ and
+  $\bm{\Tm}\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)$. Substitution is
+  again function composition, and we have $\Gamma \bm{\ext} A = (\gamma :
+  \Gamma)\times \El\,(A\,\gamma)$.
+\item Type lifting is given by $\bm{\Lift}\,p\,A = \lambda\,\gamma.\,
+  \Lift\,(p\,\gamma)\,(A\,\gamma)$.  Similarly as in the TTGU model,
+  $\bm{\Tm}\,\Gamma\,A = \bm{\Tm}\,\Gamma\,(\bm{\Lift}\,p\,A)$ and $\Gamma \bm{\ext}
+  A = \Gamma \bm{\ext} (\bm{\Lift}\,p\,A)$ follow from the $\msf{ElLift}$
+  equality, and term lifting is the identity function.
+\item We define $\bm{\U}\,i\,j\,p = \lambda\,\gamma.\,\U'\,(p\,\gamma)$. Again,
+  we have $\bm{\Tm}\,\Gamma\,(\U\,i\,j\,p) = \bm{\Ty}\,\Gamma\,i$ by Lemma \ref{lem:ucomp},
+  and $\bm{\El}$ and $\bm{\Code}$ are identity functions.
+\item Basic type formers are interpreted using $\uir$ codes. Preservation of
+  type and term formers by lifting follows by the definition of $\Lift$ and
+  $\El$.
+\end{enumerate}
+\end{mydefinition}
+
+To summarize, the only interesting change compared to the TTGU model is that
+levels and level morphisms gain potential dependency on contexts. However, in the
+IR model, this is simply the addition of an extra semantic function parameter.
+
+\subsection{Level Reflection}\label{sec:level_reflection}
+
+\begin{mydefinition}[Level reflection rules]
+Assume that we have \emph{definitions for internal levels} in the syntax of TTDL (i.e.\
+the initial TTDL model):
+\begin{alignat*}{3}
+  &\Lvl^I               &&: \Ty\,\Gamma\,l_0\\
+  &l_0^I,\,l_1^I         &&: \Tm\,\Gamma\,\Lvl^I \\
+  &\blank\!<^I\!\blank  &&: \Tm\,\Gamma\,\Lvl^I \to \Tm\,\Gamma\,\Lvl^I \to \Ty\,\Gamma\,l_0\\
+  &l_{01}^{I}            &&: \Tm\,\Gamma\,(l_0^I <^I l_1^I)
+\end{alignat*}
+A \emph{reflection rule} for the above consists of
+\begin{enumerate}
+  \item $\mkLvl : \Tm\,\Gamma\,\Lvl^I \to \Lvl\,\Gamma$ with its inverse $\unLvl$, such that $\mkLvl\,l_0^I = l_0$ and $\mkLvl\,l_1^I = l_1$.
+  \item $\mkMor\,\,: \Tm\,\Gamma\,(i <^I j) \to \mkLvl\,i < \mkLvl\,i$ with its inverse
+        $\unMor$.
+\end{enumerate}
+\end{mydefinition}
+
+For any definition of internal levels, we may extend the specification of TTDL
+with the corresponding reflection rule, thereby getting an algebraic signature
+for a TTFL. We can easily get a TTFL with an IR model in the following
+way. First, pick an external $\Lvl$ structure which a) satisfies the
+bootstrapping assumption b) has sets of levels and morphisms which can be
+represented with syntactic TTDL types.
+
+For example, if $\Lvl$ = ($\Nat$, $\blank\!<\!\blank$), with $l_0 = 0$ and $l_1
+= 1$, and TTDL supports natural numbers, then we can define $\Lvl^I$ as the
+internal $\Nat_{l_0}$, and define $\blank\!<^I\!\blank$ as the usual ordering of
+numbers, using TTDL type formers and large elimination (which is available from
+$l_0 < l_1$). Then it follows that the model in Definition \ref{def:ttdlmodel},
+instantiated to the current level structure, satisfies level reflection. The
+model even supports the stricter $\Tm\,\Gamma\,\Nat_{l_0} = \Lvl\,\Gamma$
+equation, but in general it is easier to set up models if only an isomorphism
+is required.
+
+\section{Universe Features in TTFL}
+
+In this section we give a tour of some of the features expressible in TTFL.
+
+\textbf{Bounded universe polymorphism} is realized by quantifying
+over levels and morphisms with the usual $\Pi$ types. For example, if internal
+levels strictly correspond to natural numbers, we may have
+\begin{alignat*}{3}
+  &\msf{idUpTo3} : \Pi(l : \Nat_3)(p : l <^I 3)(A : \U\,l\,3\,(\mkMor\,p)) \to \Lift\,(\mkMor\,p)\,A \to \Lift\,(\mkMor\,p)\,A\\
+  &\msf{idUpTo3} = \lambda\,l\,p\,A\,a.\,a
+\end{alignat*}
+Here, we make sure that all types in the $\Pi$ are on the same level. We assume
+that internal levels are in $\Nat_0$, but we can bind an $l : \Nat_3$, because
+by cumulativity $l$ is also a term of $\Nat_0$.
+
+\textbf{Transfinite hierarchies} are naturally supported. For example, $\Lvl$
+can be identified with $\msf{Maybe}\,\Nat_0$, where $\msf{Nothing}$ defines
+$\omega$ and $\msf{Just}\,n$ is a finite level. Then, by the definition of
+morphisms, we have $<\!\omega : \Pi(n : \Nat_0) \to \msf{Just}\,n <^I \omega$.
+We can use this to quantify over finite levels, as in the following type:
+\begin{alignat*}{3}
+  \Pi(n : \Nat_{\omega})(A : \U\,n\,\omega\,(\mkMor\,(<\!\omega\,n))) \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A
+\end{alignat*}
+
+\textbf{Induction on levels and level morphisms.} In Agda 2.6.1, there is an
+internal type of finite levels, and while construction rules and some built-in
+operations on levels are exposed, there is no general elimination rule on levels.
+Thus, there is a $\Nat \to \Lvl$ function but no non-constant $\Lvl \to \Nat$ one.
+In contrast, TTFL supports arbitrary elimination on levels, and in particular levels
+can be definitionally isomorphic to natural numbers.
+
+\textbf{Type formers returning in least upper bounds of levels}. It is common in
+type theories to allow type formers to have parameter types in different
+universe levels, say $i$ and $j$, and return in level $i \sqcup j$. In TTFL,
+whenever levels are \emph{trichotomous}, meaning that the ordering and equality of
+levels is internally decidable, then $\sqcup$ is a definable operation, and the
+``heterogeneous'' type formers are derivable\footnote{A level structure which is trichotonomous and supports \emph{extensionality}, i.e.\ $(\forall i.\, (i < j) \iff (i < k)) \to j = k$, is called a \emph{type-theoretic ordinal}.
+Assuming excluded middle type-theoretic ordinals are equivalent to classical ordinals \cite[Section 10.3]{hottbook}.}.
+
+\textbf{Coercive cumulative subtyping.} TTFL does not directly support
+subtyping. However, given enough type formers, it is possible to simply define
+subtyping as an interal inductive relation, and interpret a witness of $A \leq B$ as
+a coercion function from $A$ to $B$. Also, it should be possible to model a
+theory with cumulative subtyping by translation to coercions in TTFL.
+
+  %% data _≤_ (i : Lvl) (j : Lvl) : U i → U j → U (
+  %% U≤ : (p : i ≤ j) → U i ≤ U j
+  %% Π≤ : A ≤
+
+%% TTFL does not directly support a
+%% cumulative subtyping relation. However, it is possible to model subtyping by
+%% translation to coercions in TTFL. For example:
+%% \begin{itemize}
+%%   \item Given $p : i < j$, we may coerce terms from type $\U\,i\,k\,q$ to $\U\,j\,k\,(p \circ q)$,
+%%         using $\Lift\,p$.
+%%   \item If there is a coercion from $A$ to $A'$, and
+
+%% \end{itemize}
 

### 2021-02-21 paper.tex
progress

@@
 universe levels are first-class types and subject to arbitrary internal
-reasoning. This generalizes the bounded universe polymorphism features of Coq
-and also the internal level computations in Agda.
+reasoning. This generalizes the bounded polymorphism features of Coq and at the
+same time the internal level computations in Agda.
 \end{abstract}
@@
 
-\todo{purge unused features}
-
 We work in a Martin-Löf type theory which has the following features.
@@
     inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
-    $(t,\, u)$ and projections as $t.1$ and $t.2$. We have $\top$ as the unit
-    type with inhabitant $\tt$, and $\bot$ as the empty type. Propositional identity
-    is written as $t = u$ with $\refl : t = t$.
+    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, and
+    $\bot$ as the empty type. Propositional identity is written as $t = u$
+    (coinciding with definitional equality).
   \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
-    function types with implicit arguments, and omit implicit applications
-    whenever they can be inferred. We may omit implicit function types
-    altogether if it is clear where certain variables are quantified.
+    function types with implicit arguments. We usually omit implicit
+    applications but may explicitly write them as $t\,\{u\}$.  We may omit
+    implicit function types altogether if it is clear where certain variables
+    are quantified.
 \end{itemize}
@@
   $t[\sigma]$ for substituting types and terms. Additionally, a family structure
-  has \emph{context comprehension} which consists of an context extension
+  has \emph{context comprehension} which consists of a context extension
   operation $\blank\ext\blank : (\Gamma : \Con) \to \Ty\,\Gamma \to \Con$
@@
 There are other ways for presenting the basic categorical structure of models,
-which are nonetheless equivalent to cwfs, including natural models \cite{awodey18natural}
-and categories with attributes \cite{cartmellthesis}. We use the cwf presentation for its
-immediately algebraic character and closeness to syntax, as an explicit
-substitution calculus can be directly read off the specification.
+which are nonetheless equivalent to cwfs, including natural models
+\cite{awodey18natural} and categories with attributes \cite{cartmellthesis}. We
+use the cwf presentation for its immediately algebraic character and closeness
+to conventional explicit substitution syntax.
 
@@
 
-In the following, whenever we introduce a type former, we always assume that it
-is natural with respect to substitution, i.e.\ all type and term formers have a
-corresponding substitution rule.
+In the following, whenever we introduce a type or term former, we always assume
+that it is natural with respect to substitution, i.e.\ all type and term formers
+have a corresponding substitution rule.
 
@@
 
-\todo{Mention dependent right adjoints}
+We may think about the relation between \emph{modalities} and liftings.  The
+main difference is that we have no structural restrictions on variables and
+contexts. More concretely, every $\Lift$ is dependent right adjoint
+\cite{dependentrightadjoints} to the identity functor on the base category, as
+we have $\Tm\,(\msf{Id}\,\Gamma)\,A \simeq \Tm\,\Gamma\,(\Lift\,A)$. Hence,
+every lifting can be viewed as a degenerate modality.
 
@@
 \Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
-computes appropriately on canonical terms: if $b$ is $\true_0$, then we get
+computes appropriately on canonical terms: if $b$ is $\true_0$, we get
 $\up\!\true_0 = \true_1$ as the if-then-else scrutinee.
@@
 \cite{sterling2019algebraic}. The additional equations are a matter of convenience: they allow us
-to omit term liftings in informal syntax\footnote{In a proof assistant, we would
+to omit term liftings in informal syntax\footnote{In a proof assistant, often we would
   still have to explicitly transport along the strict inclusion
@@
 
-\begin{mylemma}
-Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
-\end{mylemma}
+%% \begin{mylemma}
+%% Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
+%% \end{mylemma}
 
@@
 to larger families.
-
-\todo{find reference for Coquand univ}
 
@@
 \begin{mydefinition}[Level reflection rules]
-Assume that we have \emph{definitions for internal levels} in the syntax of TTDL (i.e.\
-the initial TTDL model):
+Assume that we have definitions for internal levels in the syntax of TTDL,
+i.e.\ all of the following are defined:
 \begin{alignat*}{3}
@@
 with the corresponding reflection rule, thereby getting an algebraic signature
-for a TTFL. We can easily get a TTFL with an IR model in the following
-way. First, pick an external $\Lvl$ structure which a) satisfies the
-bootstrapping assumption b) has sets of levels and morphisms which can be
-represented with syntactic TTDL types.
+for a type theory with first-class levels (TTFL). We can easily get a TTFL with
+an IR model in the following way. First, pick an external $\Lvl$ structure which
+a) satisfies the bootstrapping assumption b) has sets of levels and morphisms
+which can be represented with syntactic TTDL types.
 
@@
 
-In this section we give a tour of some of the features expressible in TTFL.
+We describe some of the features expressible in TTFL.
 
@@
 internal type of finite levels, and while construction rules and some built-in
-operations on levels are exposed, there is no general elimination rule on levels.
-Thus, there is a $\Nat \to \Lvl$ function but no non-constant $\Lvl \to \Nat$ one.
-In contrast, TTFL supports arbitrary elimination on levels, and in particular levels
-can be definitionally isomorphic to natural numbers.
+operations on levels are exposed, there is no general elimination rule on
+levels. Thus, there is a $\Nat \to \Lvl$ conversion function but it has no
+inverse. In contrast, TTFL supports arbitrary elimination on levels and
+morphisms.
 
@@
 universe levels, say $i$ and $j$, and return in level $i \sqcup j$. In TTFL,
-whenever levels are \emph{trichotomous}, meaning that the ordering and equality of
-levels is internally decidable, then $\sqcup$ is a definable operation, and the
-``heterogeneous'' type formers are derivable\footnote{A level structure which is trichotonomous and supports \emph{extensionality}, i.e.\ $(\forall i.\, (i < j) \iff (i < k)) \to j = k$, is called a \emph{type-theoretic ordinal}.
-Assuming excluded middle type-theoretic ordinals are equivalent to classical ordinals \cite[Section 10.3]{hottbook}.}.
-
-\textbf{Coercive cumulative subtyping.} TTFL does not directly support
-subtyping. However, given enough type formers, it is possible to simply define
-subtyping as an interal inductive relation, and interpret a witness of $A \leq B$ as
-a coercion function from $A$ to $B$. Also, it should be possible to model a
-theory with cumulative subtyping by translation to coercions in TTFL.
-
-  %% data _≤_ (i : Lvl) (j : Lvl) : U i → U j → U (
-  %% U≤ : (p : i ≤ j) → U i ≤ U j
-  %% Π≤ : A ≤
-
-%% TTFL does not directly support a
-%% cumulative subtyping relation. However, it is possible to model subtyping by
-%% translation to coercions in TTFL. For example:
-%% \begin{itemize}
-%%   \item Given $p : i < j$, we may coerce terms from type $\U\,i\,k\,q$ to $\U\,j\,k\,(p \circ q)$,
-%%         using $\Lift\,p$.
-%%   \item If there is a coercion from $A$ to $A'$, and
-
-%% \end{itemize}
-
-%% --------------------------------------------------------------------------------
+whenever levels are \emph{trichotomous}, meaning that the ordering and equality
+of levels is internally decidable, $i \sqcup j$ can be defined as the greater of
+$i$ and $j$, and the ``heterogeneous'' type formers are derivable\footnote{A
+level structure which is trichotonomous and supports \emph{extensionality},
+i.e.\ $(\forall i.\, (i < j) \iff (i < k)) \to j = k$, is a
+\emph{type-theoretic ordinal}. Assuming excluded middle, type-theoretic
+ordinals are equivalent to classical ordinals \cite[Section 10.3]{hottbook}.}.
+
+\textbf{Coercive cumulative subtyping.}
+TTFL as specified does not directly support cumulative subtyping. However, it is
+compatible with coercive subtyping. Consider the following rules:
+\begin{alignat*}{3}
+  &\blank\!\leq\!\blank &&: \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j \to \Seti\\
+  &\msf{coerce} &&: A \leq B \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,B\\
+  &\leq\!\msf{refl} &&: A \leq A \\
+  &\U\!\leq    &&: i < i' \to \U\,i\,j\,p \leq \U\,i'\,k\,q\\
+  &\Pi\!\leq   &&: (p : A' \leq A) \to ((a' : \Tm\,\Gamma\,A')
+  \to B[x \mapsto \msf{coerce}\,p\,a'] \leq B'[x \mapsto a'])\\
+  & && \hspace{1em} \to \Pi(x : A) B \leq \Pi (x : A') B'
+\end{alignat*}
+Any model of TTFL can support the above rules: we can define
+$\blank\!\leq\!\blank$ and $\msf{coerce}$ by indexed induction-recursion
+\cite{indexedir}, where we define coercion along $\U\!\leq$ by type lifting, and
+coercion along $\Pi\!\leq$ by backwards-forwards coercion.
+
+Note that $\Pi$ is contravariant in the domain. This is easily supported
+with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
+\cite{timany2018cumulative}, where function domains are invariant.
+
+\section{Related Work} \label{sec:related}
+
+\section{Future Work and Conclusion} \label{sec:conclusion}
 

### 2021-02-21 paper.tex
progress

@@
 
-\noindent Several related topics are not handled in this paper.
-
-\begin{itemize}
-
-\item
-Besides consistency, we are often interested in \emph{canonicity},
-\emph{normalization} or other metatheoretical properties. The current work
-focuses on consistency and leaves other properties to future work. However, we
-do keep canonicity in mind while specifying our systems, and aim to avoid
-pitfalls which would cause canonicity to fail.
-\item
-We only focus on using universes as sized-based classifiers for
-types. Universe-like stratification is also present in two-level type theory
-\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
-in homotopy type theory \cite{hottbook}, but we do not attempt to cover such use
-cases.
-\item
-We only cover predicative hierarchies.
-\item
-We do not discuss ergonomics and implementation of elaboration. Our goal is to
-develop a system which is as flexible as possible, even though that may not be
-the best suited for inference and elaboration. For example, first-class levels
-have undecidable ordering in general.
-\end{itemize}
-
 We provide an Agda formalization of the contents of the paper. The formalization
@@
 \item A lifting structure with
+  %% \vspace{-0.5em}
   \begin{alignat*}{3}
@@
   Such that
-  \vspace{1em}
   \begin{enumerate}
@@
   \end{enumerate}
-  \vspace{1em}
+
   Above we mention basic type formers, although we haven't yet specified
@@
 \item A universe structure
+  %% \vspace{-0.5em}
   \begin{alignat*}{6}
@@
 
-\section{Future Work and Conclusion} \label{sec:conclusion}
+\section{Conclusion and Future Work} \label{sec:conclusion}
+
+In the current work, we developed a framework for modeling a variety of universe
+features in type theories. At this point, we may ask the question: if IR is
+sufficient to model every feature, why not simply add IR support in a practical
+implementation, and drop the menagerie of universe features?
+
+The answer is that IR provides is a \emph{deep embedding} of universe features,
+which is usually less convenient to use as \emph{native} features. For example,
+both Coq and Agda have powerful automatic solving for filling out implicit
+universe levels. We also do not have to invoke $\El$ or the $\ult$ computation
+rule explicitly, and in Coq we can use implicit syntax for subtyping instead of
+explicit coercions.
+
+This trade-off between convenience and formal minimalism is similar to the
+situation with inductive types. Formally, W-types and identity types are easier
+to handle than general inductive families, but the latter are far more
+convenient to actually use. Ideally, we would like to justify complicated
+convenience features by reduction to minimal features. With the current paper,
+we hope to have made progress in this manner.
+
+\subsection{Future Work}
+
+Several related are not discussed in this paper and could be subject to future
+work.
+
+First, besides consistency, we are often interested in \emph{canonicity},
+\emph{normalization} or other metatheoretical properties. The current work
+focuses on consistency and leaves other properties to future work. We did keep
+canonicity in mind when specifying the systems in this paper. Hopefully the
+usual proof method of gluing (in other words, proof-relevant logical predicates)
+\cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
+this paper. It seems that the main task in doing so is to define proof-relevant
+logical predicates over $\uir$.
+
+Second, we only focus on using universes as sized-based classifiers for
+types. Stratification features are also present in two-level type theory
+\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
+in homotopy type theory \cite{hottbook}. It would be interesting to port
+universe features in this paper to two-level type theory, as they would
+hopefully model a form of stage polymorphism in multi-stage compilation.
+
+Third, we do not discuss implementation strategies and ergonomics of universe
+features. Which universe hierarchies support good proof automation? What kind of
+impact do first-class levels have on elaboration algorithms? Hopefully the
+current work can aid answering these questions, by at least giving a way to quickly
+check if some features are logically consistent.
+
+Lastly, we do not handle impredicative universes. The main reason for this is
+that we do not know the consistency of having IR and impredicative function
+space together in the same universe, and modeling impredicativity seems to
+require this assumption in the metatheory. This could be investigated as well
+in future research.
 

### 2021-02-21 paper.tex
progress

@@
 
-\section{Universe Features in TTFL}
+\subsection{Universe Features in TTFL}
 
@@
 \end{alignat*}
+This type is in $\Ty\,\Gamma\,\omega$, but it is not in any universe, since
+$\omega$ is the greatest level.
 
@@
 
+\subsection{Effects of Choice of Level Structure}
+
+TTFL features clearly vary depending on level structures. We make some basic
+observations.
+
+\begin{itemize}
+\item We did not mandate that the level of $\Lvl^I$ is the least level,
+i.e.\ that $l_0 < i$ for every $i \neq l_0$. If this holds, then it is
+possible to have level polymorphism at every level: at $l_0$ we can just bind
+a $\Lvl^I$, and at every other level, we can lift $\Lvl^I$ to that
+level. However, levels are not necessarily totally ordered, and $l_0$ does not
+have to be the least. This means that universe polymorphism is prohibited in
+levels which are not connected to $l_0$.
+
+\item If levels are given by a limit ordinal, then every TTFL type is contained
+in a universe. If levels form a successor ordinal, then this is not the
+case. For example, Agda 2.6.1 has $\omega + 1$ levels (externally), where
+$\Set{\omega}$ is the topmost universe, but $\Set{\omega}$ has no type.
+
+\item
+While it is possible to quantify over all levels (using plain $\Pi$ types), it
+is not possible to have level polymorphism over all levels. We may try to type
+an identity function for all levels, as $\Pi(i : \Lift\,?\,\Lvl^I)(A :
+\U\,(\mkLvl\,i)\,?\,?) \to \Lift\,?\,A \to \Lift\,?\,A$. The issue is in
+$\U\,(\mkLvl\,i)\,?\,?$, where we have to find a level which is larger than
+\emph{every} level. The solution to this issue is to simply add more levels. For
+example, for polymorphism over finite levels, we may pick $\omega + \omega$ as
+the first limit ordinal which can internally type finite level polymorphism;
+this is what Agda 2.6.2 does.
+\end{itemize}
+
 \section{Related Work} \label{sec:related}
+
+Universes:
+Martin-Löf (U in U), Girard, Martin-Löf (countable predicative),
+CoC, Coq, Luo ECC. Universe poly in Agda (?), in Coq, cumulativity,
+Sterling, Coquand normalization (?)
+
+Transfinite: super, higher-order, Mahlo, IR, indexed IR
+
+Harper univ checking?
 

### 2021-02-22 paper.tex
progress

@@
 features:
-
 \begin{itemize}
-\item Agda recently added a limited version cumulativity as an optional feature
+\item Agda recently added a limited cumulativity as an optional feature
   for universes \cite{agdadocs}, and the upcoming 2.6.2 version will extend the $\omega+1$
   universe hierarchy to $\omega*2$.
-\item Coq added support for cumulative inductive types \cite{timany18cumulative}, and added a form of
-  bounded universe polymorphism \cite{ziliani15unification}.
+\item Coq added support for cumulative inductive types \cite{timany18cumulative}
+  and a form of bounded universe polymorphism \cite{ziliani15unification}.
 \end{itemize}
-
 \noindent At this point, there is a veritable zoo of universe features in existing
@@
   \item We first give a specification of \emph{type theory with dependent
-        levels}, or TTDL, where levels and level morphism may depend on typing contexts.
+        levels}, or TTDL, where levels and level morphisms may depend on typing contexts.
         Here, liftings, universes and type formers are specified, but the internal
@@
 \U\,(\mkLvl\,i)\,?\,?) \to \Lift\,?\,A \to \Lift\,?\,A$. The issue is in
-$\U\,(\mkLvl\,i)\,?\,?$, where we have to find a level which is larger than
-\emph{every} level. The solution to this issue is to simply add more levels. For
-example, for polymorphism over finite levels, we may pick $\omega + \omega$ as
-the first limit ordinal which can internally type finite level polymorphism;
-this is what Agda 2.6.2 does.
+$\U\,(\mkLvl\,i)\,?\,?$, where we would have to find a level which is larger
+than \emph{every} level. The solution to this issue is to simply add more
+levels. For example, for polymorphism over finite levels, we may pick $\omega +
+\omega$ as the first limit ordinal which can internalize finite level
+polymorphism; this is what Agda 2.6.2 does.
 \end{itemize}
@@
 
-Universes:
-Martin-Löf (U in U), Girard, Martin-Löf (countable predicative),
-CoC, Coq, Luo ECC. Universe poly in Agda (?), in Coq, cumulativity,
-Sterling, Coquand normalization (?)
-
-Transfinite: super, higher-order, Mahlo, IR, indexed IR
-
-Harper univ checking?
+Predicative hierarchies originate as Russell's ramified type theories
+\cite{principia}. In the more modern formulations of type theory, Martin-Löf
+proposed a countable predicative hierarchy \cite{martinlof73predicative}, as a
+way to remedy the inconsistency of the previous version of the theory (which
+assumed type-in-type). Harper and Pollack described universe inference with
+level assignments and also a form of level polymorphism \cite{harperpollack}.
+Sterling \cite{sterling2019algebraic} gave an algebraic specification much like
+ours for a type theory with countable cumulative universes, and prove canonicity
+for it.
+
+There have been proposals for strengthening universes with various closure
+principles and universe operators. Palmgren's super universes and higher-order
+universes \cite{Palmgren98onuniverses} and Setzer's Mahlo universes
+\cite{setzer00mahlo} are examples for this. These are sufficient to model
+transfinite hierarchies, but as we noted in Section
+\ref{sec:inductive_recursive_codes}, we do not know how to model strict
+inclusions with them. Variants of induction-recursion \cite{dybjer99finite,
+  indexedir, positiveir} are particularly flexible and powerful extensions to
+universes.
+
+It is worth to summarize here the universe features in the current
+type theory implementations.
+
+\textbf{Agda 2.6.1} has $\omega+1$-many non-cumulative predicative universes as
+$\Set{i}$, with optional cumulative subtyping only for universes (and not other
+type formers) \cite{agdadocs}.  It also has an internal type $\msf{Level} : \Set0$ for finite
+levels (hence, exluding $\omega$), which supports constructors and some built-in
+operations, but no general elimination rule. There is also an $\omega$-bounded
+parallel hierarchy $\msf{Prop_i}$ for strict propositions \cite{sprop}. Agda
+2.6.2 will extend the hierarchy to $\omega * 2$.
+
+\textbf{Coq 8.13} has $\omega$-many cumulative predicative universes with
+cumulative subtyping for all type formers \cite{timany18cumulative}. It supports
+bounded universe polymorphism, but it has no internal type for levels, and
+universe polymorphic definition are not internally typeable. It also optionally
+has an impredicative $\msf{Prop}$ universe and optionally impredicative bottom
+$\Seti$ universe. Version 8.13 added experimental support for a parallel
+$\omega$-bounded cumulative hierarchy for strict propositions.
+
+\textbf{Lean 3.3} has countable non-cumulative predicative $\msf{Type}_i$
+universes with universe polymorphism, and no internal type of levels
+\cite{leanmanual}. It also has strict impredicative $\msf{Prop}$.
+
+\textbf{Idris 1} has countable cumulative universes with cumulative subtyping
+only for universes, typical-ambiguity-style level inference and no universe
+polymorphism \cite{idrisdocs}.
+
+Of the above features, what TTFL does not support is a) impredicativity b) the
+interaction of $\msf{Prop}$ and $\msf{Type}$ universes, i.e. the restrictions on
+$\msf{Prop}$ elimination.
 
@@
 universe features in this paper to two-level type theory, as they would
-hopefully model a form of stage polymorphism in multi-stage compilation.
+hopefully model a form of stage polymorphism in multi-stage compilation. It
+would be nice to try representing $\msf{Prop}$ universes in TTFL as well;
+this is closed related to h-level based stratification.
 

### 2021-02-22 paper.tex
progress

@@
 
-Users of type theories often view universe levels as an unwieldy bureaucratic
-detail, a necessary annoyance in service of boosting expressive power while
-retaining logical consistency. However, universe hierarchies are not going away
-any time soon in practical implementations of type theory. In recent
-developments of systems, we are getting more universes and more adjacent
-features:
+Users of type theories often view universe levels as a bureaucratic detail, a
+necessary annoyance in service of boosting expressive power while retaining
+logical consistency. However, universe hierarchies are not going away any time
+soon in practical implementations of type theory. In recent developments of
+systems, we are getting more universes and more adjacent features:
 \begin{itemize}
@@
 levels? What kind of universe polymorphism do we have, can we quantify over
-bounds? Is there a type of levels, or are levels in a separate syntactic layer?
+level bounds? Is there a type of levels, or are levels in a separate syntactic layer?
 
@@
     inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
-    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, and
-    $\bot$ as the empty type. Propositional identity is written as $t = u$
-    (coinciding with definitional equality).
+    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, $\bot$ as
+    the empty type, and $\Bool$ with $\true$ and $\false$
+    inhabitants. Propositional identity is written as $t = u$ (coinciding with
+    definitional equality).
   \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
@@
   typing contexts, morphisms to parallel substitutions and the terminal object to
-  the empty context; this informs the naming scheme that we use here.
+  the empty context; this informs the naming scheme.
 \item A \emph{family structure}, containing $\Ty : \Con \to \Seti$ and $\Tm :
@@
 on ``sizing'' hierarchies instead of staging hierarchies. This means that we
-want to eliminate from any family to any other family which is ``connected'' by
+want to eliminate from any family to any other family which is connected by
 a morphism.
@@
 lifting and term lifting, but while term lifting is invertible, type lifting is
-not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and a larger $\Ty$ is
-not isomorphic a smaller one, because it contains more universes. So, for
-example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to $\Ty_1\,\Gamma$ yields
-$\Bool_1$, but lifting $\Bool_0$ as a term yields $\Bool_0$.
+not. $\Lift$ sends a $\Ty_i\,\Gamma$ to a $\Ty_j\,\Gamma$, and $\Ty_j\,\Gamma$
+is not isomorphic to $\Ty_i\,\Gamma$, because it contains more universes. So,
+for example, lifting $\Bool_0 : \Ty_0\,\Gamma$ as a type to $\Ty_1\,\Gamma$
+yields $\Bool_1$, but lifting $\Bool_0$ as a term yields $\Bool_0$.
 
@@
 
-\textbf{Bounded universe polymorphism} is realized by quantifying
-over levels and morphisms with the usual $\Pi$ types. For example, if internal
-levels strictly correspond to natural numbers, we may have
-\begin{alignat*}{3}
-  &\msf{idUpTo3} : \Pi(l : \Nat_3)(p : l <^I 3)(A : \U\,l\,3\,(\mkMor\,p)) \to \Lift\,(\mkMor\,p)\,A \to \Lift\,(\mkMor\,p)\,A\\
+\textbf{Bounded universe polymorphism} is realized by quantifying over levels
+and morphisms with the usual $\Pi$ types. For example, if levels strictly
+correspond to internal natural numbers, we may have
+\begin{alignat*}{3}
+  &\msf{idUpTo3} : \Pi(l : \Nat_3)(p : \Lift_0^3(l <^I 3))(A : \U\,l\,3\,(\mkMor\,p)) \to \Lift\,(\mkMor\,p)\,A \to \Lift\,(\mkMor\,p)\,A\\
   &\msf{idUpTo3} = \lambda\,l\,p\,A\,a.\,a
 \end{alignat*}
-Here, we make sure that all types in the $\Pi$ are on the same level. We assume
-that internal levels are in $\Nat_0$, but we can bind an $l : \Nat_3$, because
-by cumulativity $l$ is also a term of $\Nat_0$.
+Here, we make sure that all types are on the same level, by appropriate
+lifting. We assume that internal levels are in $\Nat_0$, but we can bind an $l :
+\Nat_3$, because by cumulativity $l$ is also a term of $\Nat_0$. Likewise, the
+$p$ variable is a term of $\Lift_0^3(l <^I 3)$ and $l <^I 3$ as well.
 
@@
 with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
-\cite{timany2018cumulative}, where function domains are invariant.
+\cite{timany18cumulative}, where function domains are invariant.
 
@@
 
-Several related are not discussed in this paper and could be subject to future
-work.
+Several related topics are not discussed in this paper and could be subject to
+future work.
 
@@
 \cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
-this paper. It seems that the main task in doing so is to define proof-relevant
-logical predicates over $\uir$.
+this paper.
 
@@
 universe features in this paper to two-level type theory, as they would
-hopefully model a form of stage polymorphism in multi-stage compilation. It
-would be nice to try representing $\msf{Prop}$ universes in TTFL as well;
-this is closed related to h-level based stratification.
+hopefully model a form of stage polymorphism in multi-stage compilation. We
+could try representing $\msf{Prop}$ universes in TTFL as well. This is closely
+related to h-level based stratification.
 
@@
 require this assumption in the metatheory. This could be investigated as well
-in future research.
+in future work.
 

### 2021-02-22 paper.tex
Tidy up Agda, paper typos

@@
 
-Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_0$). Now, we can
+Assume an inclusion from ($\Ty_0$, $\Tm_0$) to ($\Ty_1$, $\Tm_1$). Now, we can
 eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
 \Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
-\Tm_1\,\Gamma\,(\Lift_1\,\Bool_0)$, hence $\up\!b :
+\Tm_1\,\Gamma\,(\Lift\,\Bool_0)$, hence $\up\!b :
 \Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
@@
 \cite{indexedir}, where we define coercion along $\U\!\leq$ by type lifting, and
-coercion along $\Pi\!\leq$ by backwards-forwards coercion.
+coercion along $\Pi\!\leq$ by backwards-forwards coercion. It is possible to
+extend the subtyping relation with rules for other basic type formers.
 
@@
 \textbf{Agda 2.6.1} has $\omega+1$-many non-cumulative predicative universes as
-$\Set{i}$, with optional cumulative subtyping only for universes (and not other
-type formers) \cite{agdadocs}.  It also has an internal type $\msf{Level} : \Set0$ for finite
+$\Set{i}$, with optional cumulative subtyping only for universes
+\cite{agdadocs}.  It also has an internal type $\msf{Level} : \Set0$ for finite
 levels (hence, exluding $\omega$), which supports constructors and some built-in
-operations, but no general elimination rule. There is also an $\omega$-bounded
-parallel hierarchy $\msf{Prop_i}$ for strict propositions \cite{sprop}. Agda
-2.6.2 will extend the hierarchy to $\omega * 2$.
+operations, but no general elimination rule. There is also a countable parallel
+hierarchy $\msf{Prop_i}$ for strict propositions \cite{sprop}. Agda 2.6.2 will
+extend the $\Set{i}$ hierarchy to $\omega * 2$.
 
@@
 bounded universe polymorphism, but it has no internal type for levels, and
-universe polymorphic definition are not internally typeable. It also optionally
-has an impredicative $\msf{Prop}$ universe and optionally impredicative bottom
-$\Seti$ universe. Version 8.13 added experimental support for a parallel
-$\omega$-bounded cumulative hierarchy for strict propositions.
+universe polymorphic definitions are not internally typeable. It also has an
+impredicative $\msf{Prop}$ universe and optionally impredicative bottom $\Seti$
+universe. Version 8.13 added experimental support for a parallel
+countable cumulative hierarchy for strict propositions.
 
@@
 
-\textbf{Idris 1} has countable cumulative universes with cumulative subtyping
-only for universes, typical-ambiguity-style level inference and no universe
-polymorphism \cite{idrisdocs}.
+\textbf{Idris 1} has countable cumulative predicative universes with cumulative
+subtyping only for universes, typical-ambiguity-style level inference and no
+universe polymorphism \cite{idrisdocs}.
 

### 2021-02-22 paper.tex
typo

@@
 
-Predicative hierarchies originate as Russell's ramified type theories
+Predicative hierarchies originate from Russell's ramified type theories
 \cite{principia}. In the more modern formulations of type theory, Martin-Löf
@@
 The answer is that IR provides is a \emph{deep embedding} of universe features,
-which is usually less convenient to use as \emph{native} features. For example,
+which is usually less convenient to use than \emph{native} features. For example,
 both Coq and Agda have powerful automatic solving for filling out implicit

### 2021-02-23 paper.tex
correction

@@
 
-Lifted types correspond to types of object-level expressions; for
-example, $\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
+Lifted types correspond to types of object-level expressions; for example,
+$\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
 $\Lift\,\Bool_0$ is the meta-level type of $\Bool_0$-expressions, and $\Bool_1 :
 \Ty_1\,\Gamma$ is the type of meta-level Booleans. It is possible to compute a
-$\Lift\,\Bool_0$ from a $\Bool_1$. Given $b : \Tm_1\,\Gamma\,\Bool_1$, we can
-construct
+$\Bool_0$ from a $\Bool_1$. Given $b : \Tm_1\,\Gamma\,\Bool_1$, we can construct
 $\down(\mathsf{if}\,b\,\mathsf{then}\up\!\true_0\,\mathsf{else}\up\!\false_0) :
@@
 
-But there is no way to compute a $\Bool_1$ from a $\Lift\,\Bool_0$, since
-$\Lift\,\Bool_0$ has no elimination rule in $\Ty_1$. Conceptually, terms
-of $\Lift\,\Bool_0$ are \emph{expressions}, so they are not necessarily $\true$ or
-$\false$, they can also be variables or neutral expressions, so the usual
-Boolean elimination is not justified for them. In general, the setup ensures that
-we can eliminate from positive types in $\Ty_1$ to their counterparts in
-$\Ty_0$, but not the other way around, while negative types are preserved by
-$\Lift$ up to type isomorphism.
+But there is no way to compute a $\Bool_1$ from a $\Bool_0$: we can try to lift
+the input, but there is no elimination rule for $\Lift\,\Bool_0$ in
+$\Ty_1$. Conceptually, terms of $\Lift\,\Bool_0$ are \emph{expressions}, so they
+are not necessarily $\true$ or $\false$, they can also be variables or neutral
+expressions, so the usual Boolean elimination is not justified for them. In
+general, the setup ensures that we can eliminate from positive types in $\Ty_1$
+to their counterparts in $\Ty_0$, but not the other way around, while negative
+types are preserved by $\Lift$ up to type isomorphism.
 

### 2021-02-23 paper.tex
correction

@@
 
-\subsection{Morphisms, Liftings and Inclusions of Families}
+\subsection{Morphisms and Inclusions of Families}
 \label{sec:morphisms}
@@
 \cite{kaposi2019gluing}) of cwfs: a weak morphism which has the identity action
-on the base category is exactly a family morphism. A plain family morphism is
-still not too interesting, so we add more features.
-
-\begin{mydefinition}
-A \emph{lifting} is a family morphism which is bijective on terms. Note that
-we can drop the $\ext$-preservation condition from the specification of
-liftings, since it follows from the invertible action on terms.
-\end{mydefinition}
+on the base category is exactly a family morphism.
+
+\begin{mylemma} Every family morphism has invertible action on terms, i.e.\ there
+is an $F^{-1} : \Tm\,\Gamma\,(F\,A) \to \Tm\,\Gamma\,A$. \emph{Proof:} from the
+preservation of context extension by $F$, we know that $F$ is invertible on the
+zero De Bruijn index, so for $\q : \Tm\,(\Gamma \ext F\,A)\,(F (A[\p]))$, we
+have $F^{-1}\q : \Tm\,(\Gamma \ext F\,A)\,(A[\p])$, such that $F (F^{-1}\,\q) =
+\q$. If we have any $t : \Tm\,\Gamma\,(F\,A)$, we also have $(F^{-1}\q)[\id,\,t]
+: \Tm\,\Gamma\,A$, and this yields the inverse action $F^{-1}$ which acts on all
+terms. \qed
+\end{mylemma}
 
 \begin{notation}
-We write $\Lift : \Ty_0\,\Gamma \to \Ty_1\,\Gamma$ for the action of some lifting
-on types, $\up : \Tm_0\,\Gamma\,A \to \Tm_1\,\Gamma\,(\Lift\,A)$ for the action
-on terms, and $\down$ for the inverse action on terms.
+In the following, we will write $\Lift : \Ty_0\,\Gamma \to \Ty_1\,\Gamma$ for
+the action of some morphism on types, $\up : \Tm_0\,\Gamma\,A \to
+\Tm_1\,\Gamma\,(\Lift\,A)$ for the action on terms, and $\down$ for the inverse
+action on terms. We will also call the action on types \emph{type lifting}
+and the action on terms \emph{term lifting}.
 \end{notation}
 
-We may think about the relation between \emph{modalities} and liftings.  The
+We may think about the relation between \emph{modalities} and morphisms. The
 main difference is that we have no structural restrictions on variables and
@@
 we have $\Tm\,(\msf{Id}\,\Gamma)\,A \simeq \Tm\,\Gamma\,(\Lift\,A)$. Hence,
-every lifting can be viewed as a degenerate modality.
+every morphism can be viewed as a degenerate modality.
 
@@
 Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
-lifting between them. This corresponds to a basic version of \emph{two-level
+morphism between them. This corresponds to a basic version of \emph{two-level
   type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
@@
 generative two-stage compilation with dependent types. Comparing this system to
-e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can relate $\Lift$ to $\mathsf{code}$, $\up$
-to the quasi-quotation operation $<\!\blank\!>$, and $\down$ to escape
-$\sim\!\blank$.
+e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can relate $\Lift$ to
+$\mathsf{code}$, $\up$ to the quasi-quotation operation $<\!\blank\!>$, and
+$\down$ to escape $\sim\!\blank$.
 
@@
 \begin{mydefinition}\label{def:inclusion}
-A \emph{family inclusion} is a lifting which preserves all type and term
+A \emph{family inclusion} is a family morphism which preserves all type and term
 formers. This assumes that every type former which is contained in the source
@@
 \end{alignat*}
-In general, we can skip specifying preservation for $\down$, since it follows from $\up$ preservation
-equations.
+In general, we can skip specifying preservation for $\down$, since it follows
+from $\up$ preservation equations.
 
@@
 A strict inclusion corresponds to Sterling's \emph{algebraic cumulativity}
-\cite{sterling2019algebraic}. The additional equations are a matter of convenience: they allow us
-to omit term liftings in informal syntax\footnote{In a proof assistant, often we would
-  still have to explicitly transport along the strict inclusion
-  equations.}. Most of the time we can also omit level annotations
-on term formers. For example, we have $\true_0 : \Tm_0\,\Gamma\,\Bool_0$,
-but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$, hence $\true_0 :
-\Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally equal to
-$\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using simply $\true$
-is fine whenever the family is clear from context.
+\cite{sterling2019algebraic}. The additional equations are a matter of
+convenience: they allow us to omit term liftings in informal syntax\footnote{In
+  a proof assistant, often we would still have to explicitly transport along the
+  strict inclusion equations.}. Most of the time we can also omit level
+annotations on term formers. For example, we have $\true_0 :
+\Tm_0\,\Gamma\,\Bool_0$, but also $\true_0 : \Tm_0\,\Gamma\,(\Lift\,\Bool_0)$,
+hence $\true_0 : \Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally
+equal to $\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using
+simply $\true$ is fine whenever the family is clear from context.
 
@@
 
-%% \begin{mylemma}
-%% Every well-founded relation is irreflexive \cite[Lemma 10.4.2]{hottbook}.
-%% \end{mylemma}
-
 \begin{mydefinition} A \emph{level structure} consists of the following components:
@@
 be mapped to trivial liftings, which are not interesting in our setting. Second,
-we do not need proof-relevant level morphisms, since any parallel pair of family
-liftings gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
+we do not need proof-relevant level morphisms, since any parallel pair of
+morphisms gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
 i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
@@
 and $\Lift\,q\,A$ are in the same family, we can internally prove them
-isomorphic using function types and identity types. That said, every construction
-in this paper would still work with inverse categories as level structures.
+isomorphic using function types and identity types. That said, every
+construction in this paper would still work with inverse categories as level
+structures.
 
@@
 same canonical inhabitants as $\Bool_j$, for any $i$ and $j$. Universes
-introduce dependency on levels, by serving as classifiers for smaller families internally
-to larger families.
+introduce dependency on levels, by serving as classifiers for smaller families
+internally to larger families.
 

### 2021-02-27 arxivsub/preprint.tex
update arxiv version

--- 2021-02-23 paper.tex
@@
+\pdfoutput=1
 \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
@@
 
-%\nolinenumbers %uncomment to disable line numbering
-
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+\nolinenumbers %uncomment to disable line numbering
+
+\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
 
@@
 
-\usepackage{xcolor}
-\usepackage{mathpartir}
-\usepackage{todonotes}
-\presetkeys{todonotes}{inline}{}
 \usepackage{scalerel}
-\usepackage{amssymb}
 \usepackage{bm}
-
-%% \theoremstyle{definition}
-%% \newtheorem{definition}{Definition}
-%% \newtheorem{example}{Example}
-
-%% \theoremstyle{theorem}
-%% \newtheorem{theorem}{Theorem}
-%% \newtheorem{lemma}{Lemma}
+\usepackage{hyperref}
 
@@
 
-We provide an Agda formalization of the contents of the paper. The formalization
-is not complete, as we skip proofs involving an excessive number of equality
-coercions (which are more suited to informal reasoning, using equality
-reflection), and instead focus on the key points.
+We provide an Agda formalization of the contents of the paper, at
+\url{https://github.com/AndrasKovacs/universes/tree/master/agda}. The
+formalization is not complete, as we skip proofs involving an excessive number
+of equality coercions (which are more suited to informal reasoning, using
+equality reflection), and instead focus on the key points.
 
@@
 We may think about the relation between \emph{modalities} and morphisms. The
-main difference is that we have no structural restrictions on variables and
-contexts. More concretely, every $\Lift$ is dependent right adjoint
+main difference is that morphisms impose no structural restrictions on variables
+and contexts. More concretely, every $\Lift$ is dependent right adjoint
 \cite{dependentrightadjoints} to the identity functor on the base category, as

### 2021-03-18 paper.tex
last edit
typo

@@
 
-Second, we only focus on using universes as sized-based classifiers for
+Second, we only focus on using universes as size-based classifiers for
 types. Stratification features are also present in two-level type theory

### 2021-04-06 efop_prez/prez.tex
stuff

--- 2021-02-15 department_prez/prez.tex
@@
+%% build: latexmk -pdf -pvc -xelatex prez.tex
+
 \documentclass[dvipsnames,10pt,aspectratio=169]{beamer}
@@
 \title{Universes In Type Theory}
-\date{January 29, 2021, PLC Department Workshop}
-\author{\vspace{-1em}András Kovács\footnote{The author was supported by the European Union, co-financed by the
-European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}
-\institute{Eötvös Loránd University, Budapest}
+\date{April 8, 2021, EFOP Workshop}
+\author{\normalsize{\vspace{-1em}\textbf{András Kovács}\footnote{The author was supported by the European Union, co-financed by the
+European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}}
+%% \institute{Eötvös Loránd University, Budapest}
 \begin{document}
@@
 
-We can assume the existence of a very large set such that all of ZF can fit there,
-which is called a \textbf{universe}. Then, just quantify over elements of the
-universe. The resulting theory is stronger than ZF.
+\textbf{Universe:} a set which includes every ZF set.
 \vspace{1em}
 
-If we want quantify over sets which \textbf{don't} fit (or universes
-themselves), we can just assume even more universes.
+Gödel: there is no universe inside ZF.
+\vspace{1em}
 
+We can assume more universes if we want to quantify over more sets.
 \end{frame}
@@
 
+Type theories:
+\begin{itemize}
+  \item Alternative to set theory as mathematical foundation.
+  \item Used in most general-purpose theorem provers.
+  \item Can be used for proving and programming at the same time.
+\end{itemize}
+
+Universes come up in type theories as well.
+\vspace{1em}
+
+Every theorem prover has different universe-related features.
+\vspace{1em}
+
+We want to give a general semantics covering many different features.
+
+\end{frame}
+
+\begin{frame}[fragile]{Example}
+
 Agda:
 \begin{verbatim}
-  Bool          : Set₀
+  ℕ             : Set₀
   Set₀          : Set₁
   Set₁          : Set₂
-  (Bool → Set₀) : Set₁
+  (ℕ → Set₀)    : Set₁
   (Set₀ → Bool) : Set₁
@@
 
-Assuming \texttt{Set₀ : Set₀} implies contradiction (by Russell-like argument).
+Identity function at all small sets:
+\begin{verbatim}
+  id : (A : Set₀) → A → A
+  id A x = x
+\end{verbatim}
+
+We have $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$ for all $i$.
 \vspace{1em}
 
-{\small(But type theory with \texttt{Set₀ : Set₀} is still a perfectly good Turing-complete
-programming language)}
+Assuming $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$ implies contradiction (by Russell-like argument).
 \vspace{1em}
@@
 
-\begin{frame}[fragile]{Universes in type theory}
+\begin{frame}[fragile]{Extensions}
 
@@
 \begin{verbatim}
-  myId : (l : Level) → l < 3 → (A : Set l) → A → A
-  myId l p A x = x
+  myId : (i : ℕ) → i < 3 → (A : Set i) → A → A
+  myId i p A x = x
 \end{verbatim}
@@
         and not get randomly stuck.
-  \item Is type checking decidable? (I don't plan to cover this in research).
+  \item Is proof checking decidable? (Not covered in current work).
 \end{itemize}
@@
 
-\begin{frame}{Results}
+\begin{frame}[fragile]{Features covered by the framework}
 
-Framework covers:
-\begin{enumerate}
-  \item Universe levels may come from any well-ordered set.
-  \item Polymorphism with bounds is allowed.
-  \item Universes are cumulative.
-  \item Levels can be manipulated arbitrarily as program data (WIP).
-\end{enumerate}
-
-Things we need to assume in order to prove the framework sensible:
-\begin{enumerate}
-  \item \textbf{Two} universes.
-  \item Types of certain infinitely branching trees (W-types).
-\end{enumerate}
-
-Potential further applications (in future work):
-\begin{itemize}
-  \item Information flow type systems (``secure'' and ``public'' levels).
-  \item Staged compilation (``runtime'' and ``compile time'' levels).
-\end{itemize}
+Universe levels may come from \textbf{any well-ordered set}, even transfinite:
+\begin{verbatim}
+  Set i : Set (i + 1) : ... : Set ω : Set (ω + 1) : ...
+\end{verbatim}
+\textbf{Quantification over levels}, \textbf{arbitrary computation on levels}.
+\begin{verbatim}
+   myId : (i : ℕ) → (j : ℕ) → (A : Set (i + j)) → A → A
+   myId i j A x = x
+\end{verbatim}
+\textbf{Universes are cumulative}
+\begin{verbatim}
+   ℕ : Set₀
+   ℕ : Set₁
+   ...
+\end{verbatim}
 
@@
 
-\begin{frame}
+\begin{frame}[fragile]{Implementation}
 
-This talk is a side project intended for FSCD 2021 submission (in February).
+We define a family of type theories with general universe features.
 \vspace{1em}
 
-Other 2020 publications:
-\begin{itemize}
-  \item LICS 2020, with Ambrus: ``Large and Infinitary Quotient Inductive-Inductive Types''
-  \item ICFP 2020: ``Elaboration with First-Class Implicit Function Types''
-  \item TYPES 2019 post-proceedings, with Ambrus and Ambroise Lafont: ``For
-    Finitary Induction-Induction, Induction is Enough''
-\end{itemize}
+We prove all of them consistent by reducing all fancy features to
+a single previously known feature, called \textbf{induction-recursion}.
+\vspace{1em}
+
+Induction-recursion is a type-theoretic analogue of assuming \textbf{Mahlo cardinals} in set theory.
+
+\begin{verbatim}
+   Univ   : Set
+   Nat    : Univ
+   Π      : (A : Univ) → (Interp Univ → Univ) → Univ
+
+   Interp : Univ → Set
+   Interp Nat     = ℕ
+   Interp (Π A B) = (x : Interp A) → Interp (B x)
+\end{verbatim}
+
+\end{frame}
+
+%% \begin{enumerate}
+%%   \item Universe levels may come from any well-ordered set.
+%%   \item Polymorphism with bounds is allowed.
+%%   \item Universes are cumulative.
+%%   \item Levels can be manipulated arbitrarily as program data (WIP).
+%% \end{enumerate}
+
+%% Things we need to assume in order to prove the framework sensible:
+%% \begin{enumerate}
+%%   \item \textbf{Two} universes.
+%%   \item Types of certain infinitely branching trees (W-types).
+%% \end{enumerate}
+
+%% Potential further applications (in future work):
+%% \begin{itemize}
+%%   \item Information flow type systems (``secure'' and ``public'' levels).
+%%   \item Staged compilation (``runtime'' and ``compile time'' levels).
+%% \end{itemize}
+
+%% \end{frame}
+
+\begin{frame}{Publication}
+
+``Generalized Universe Hierarchies and First-Class Universe Levels''
+\vspace{1em}
+
+Submission under review for FSCD 2021.
+
+%% This talk is a side project intended for FSCD 2021 submission (in February).
+%% \vspace{1em}
+
+%% Other 2020 publications:
+%% \begin{itemize}
+%%   \item LICS 2020, with Ambrus: ``Large and Infinitary Quotient Inductive-Inductive Types''
+%%   \item ICFP 2020: ``Elaboration with First-Class Implicit Function Types''
+%%   \item TYPES 2019 post-proceedings, with Ambrus and Ambroise Lafont: ``For
+%%     Finitary Induction-Induction, Induction is Enough''
+%% \end{itemize}
 

### 2021-04-06 paper.tex
stuff

@@
   stratification features, from two-level type theory to cumulative universes.
-\item We use induction-recursion to model the mentioned theories. We model
-  the strongest formulations for lifting and universes, namely cumulative
-  universes with Russell-style type decoding.
+\item In Section \ref{sec:semantics} we use induction-recursion to model the mentioned theories. We
+  model the strongest formulations for lifting and universes, namely cumulative universes with
+  Russell-style type decoding.
 \item In Section \ref{sec:ttfl} we describe type theories with internal types
@@
 \begin{itemize}
-  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports
-    inductive-recursive types (IR) as specified by Dybjer and Setzer
-    \cite{dybjer99finite}. The $\Set1$ universe is not essential and we only use
-    it as a convenience feature, in this paper and in the Agda formalization. We
-    may omit the universe indices if they can be inferred or if we work over
-    arbitrary indices.
+  \item Two universes named $\Set0$ and $\Set1$, where $\Set0$ supports inductive-recursive types
+    (IR) as specified by Dybjer and Setzer \cite{dybjer99finite}. We may omit the universe indices
+    if they can be inferred or if we work over arbitrary indices.
   \item Function extensionality and uniqueness of identity proofs
@@
     working in extensional type theory, to avoid noise from equality transports.
-  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$
-    inhabitants, and $\Sigma$-types as $(x : A) \times B$, with pairing as
-    $(t,\, u)$. We have $\top$ as the unit type with inhabitant $\tt$, $\bot$ as
-    the empty type, and $\Bool$ with $\true$ and $\false$
-    inhabitants. Propositional identity is written as $t = u$ (coinciding with
-    definitional equality).
+  \item We write function types as $(x : A)\to B$ with $\lambda\,x.\,t$ inhabitants. We may group
+    multiple arguments with the same type, as in $(x\,y : A) \to B$.  We have $\Sigma$-types as $(x :
+    A) \times B$, with pairing as $(t,\, u)$.  We have $\top$ as the unit type with inhabitant
+    $\tt$, $\bot$ as the empty type, and $\Bool$ with $\true$ and $\false$
+    inhabitants. Propositional identity is written as $t = u$ (coinciding with definitional
+    equality).
   \item We occasionally use $\{x : A\} \to B$ for an Agda-like notation for
@@
 \item By going right-to-left along the isomorphism, we recover \emph{substitution extension}
-      $\blank,\blank : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(A[\sigma])$. This means
+      $\blank,\blank : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Gamma\,(A[\sigma]) \to \Sub\,\Gamma\,(\Delta\ext A)$. This means
       that starting from $\epsilon$ or the identity substitution $\id$, we can iterate $\blank,\blank$
@@
 \begin{mydefinition}
-A \emph{family morphism} $F$ between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$, $\Ty_1$)
+A \emph{family morphism} $F$ between ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$)
 families consists of natural transformations mapping types to types and terms to
@@
 \begin{mylemma} Every family morphism has invertible action on terms, i.e.\ there
-is an $F^{-1} : \Tm\,\Gamma\,(F\,A) \to \Tm\,\Gamma\,A$. \emph{Proof:} from the
-preservation of context extension by $F$, we know that $F$ is invertible on the
-zero De Bruijn index, so for $\q : \Tm\,(\Gamma \ext F\,A)\,(F (A[\p]))$, we
-have $F^{-1}\q : \Tm\,(\Gamma \ext F\,A)\,(A[\p])$, such that $F (F^{-1}\,\q) =
-\q$. If we have any $t : \Tm\,\Gamma\,(F\,A)$, we also have $(F^{-1}\q)[\id,\,t]
-: \Tm\,\Gamma\,A$, and this yields the inverse action $F^{-1}$ which acts on all
-terms. \qed
+is an $F^{-1} : \Tm\,\Gamma\,(F\,A) \to \Tm\,\Gamma\,A$.
 \end{mylemma}
+\begin{proof} From the
+preservation of context extension by $F$, we know that $F$ is invertible on the zero De Bruijn
+index, so for $\q : \Tm\,(\Gamma \ext F\,A)\,(F (A[\p]))$, we have $F^{-1}\q : \Tm\,(\Gamma \ext
+F\,A)\,(A[\p])$, such that $F (F^{-1}\,\q) = \q$. If we have any $t : \Tm\,\Gamma\,(F\,A)$, we also
+have $(F^{-1}\q)[\id,\,t] : \Tm\,\Gamma\,A$, and this yields the inverse action $F^{-1}$ which acts
+on all terms.
+\end{proof}
 
@@
 
-\paragraph{Two-level type theory}
-Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
-morphism between them. This corresponds to a basic version of \emph{two-level
-  type theory} \cite{twolevel}. This theory can be interpreted as having ($\Ty_1$,
-$\Tm_1$) as a metaprogramming layer which can generate object-level programs in
-the ($\Ty_0$, $\Tm_0$) layer.
-
-Lifted types correspond to types of object-level expressions; for example,
-$\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
-$\Lift\,\Bool_0$ is the meta-level type of $\Bool_0$-expressions, and $\Bool_1 :
-\Ty_1\,\Gamma$ is the type of meta-level Booleans. It is possible to compute a
-$\Bool_0$ from a $\Bool_1$. Given $b : \Tm_1\,\Gamma\,\Bool_1$, we can construct
+Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a morphism between them. This
+corresponds to a basic version of \emph{two-level type theory} \cite{twolevel}. This theory can be
+interpreted as having ($\Ty_1$, $\Tm_1$) as a metaprogramming layer which can generate object-level
+programs in the ($\Ty_0$, $\Tm_0$) layer. Lifted types correspond to types of object-level
+expressions; for example, $\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
+$\Lift\,\Bool_0$ is the meta-level type of $\Bool_0$-expressions, and $\Bool_1 : \Ty_1\,\Gamma$ is
+the type of meta-level Booleans. It is possible to compute a $\Bool_0$ from a $\Bool_1$. Given $b :
+\Tm_1\,\Gamma\,\Bool_1$, we can construct
 $\down(\mathsf{if}\,b\,\mathsf{then}\up\!\true_0\,\mathsf{else}\up\!\false_0) :
@@
 expressions, so the usual Boolean elimination is not justified for them. In
-general, the setup ensures that we can eliminate from positive types in $\Ty_1$
+general, the setup ensures that we can map from positive types in $\Ty_1$
 to their counterparts in $\Ty_0$, but not the other way around, while negative
@@
 
-Remarkably, the simple rules of two-level type theory appear to model a form of
-generative two-stage compilation with dependent types. Comparing this system to
-e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can relate $\Lift$ to
-$\mathsf{code}$, $\up$ to the quasi-quotation operation $<\!\blank\!>$, and
+The rules of two-level type theory appear to model a form of generative two-stage compilation with
+dependent types. Comparing this system to e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can
+relate $\Lift$ to $\mathsf{code}$, $\up$ to the quasi-quotation operation $<\!\blank\!>$, and
 $\down$ to escape $\sim\!\blank$.
@@
 \begin{mydefinition}\label{def:strict_inclusion}
-A \emph{strict family inclusion} between ($\Tm_0$, $\Ty_0$) and ($\Tm_1$,
-$\Ty_1$) is a family inclusion ($\Lift$, $\up$, $\down$) for which the following equations hold:
+A \emph{strict family inclusion} between ($\Ty_0$, $\Tm_0$) and ($\Ty_1$,
+$\Tm_1$) is a family inclusion ($\Lift$, $\up$, $\down$) for which the following equations hold:
 \begin{alignat}{2}
@@
 \begin{alignat*}{3}
-  & \Acc : \{A : \Seti\}(R : A \to A \to \Seti) \to A \to \Seti \\
+  & \Acc : \{A : \Seti\}\to(R : A \to A \to \Seti) \to A \to \Seti \\
   & \acc : \{a : A\} \to ((\mi{a'} : A) \to R\,\mi{a'}\,a \to \Acc\,R\,\mi{a'}) \to \Acc\,R\,a
@@
 
-Our choice of level structures and diagrams is motivated by the following
-reasons. First, we do not need identity morphisms in levels, because they would
-be mapped to trivial liftings, which are not interesting in our setting. Second,
-we do not need proof-relevant level morphisms, since any parallel pair of
-morphisms gives rise to isomorphic types. Concretely, given $p : i < j$ and $q :
-i < j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq
-\Tm_i\,\Gamma\,A \simeq \Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$
-and $\Lift\,q\,A$ are in the same family, we can internally prove them
-isomorphic using function types and identity types. That said, every
-construction in this paper would still work with inverse categories as level
-structures.
+Our choice of level structures and diagrams is motivated by the following.  First, we do not need
+identity morphisms in levels, because they would be mapped to trivial liftings, which are not
+interesting in our setting. Second, we do not need proof-relevant level morphisms, since any
+parallel pair of morphisms gives rise to isomorphic types. Concretely, given $p : i < j$ and $q : i
+< j$ such that $p \neq q$, we have $\Tm_j\,\Gamma\,(\Lift\,p\,A) \simeq \Tm_i\,\Gamma\,A \simeq
+\Tm_j\,\Gamma\,(\Lift\,q\,A)$, and since $\Lift\,p\,A$ and $\Lift\,q\,A$ are in the same family, we
+can internally prove them isomorphic using function types and identity types. That said, every
+construction in this paper would still work with inverse categories as level structures.
 
@@
   Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$
-  and $\elir$ by simultaneous IR. For basic type formers, below we only include
-  codes for function types, the empty type, and $\Bool$; other type formers are
-  straightforward to add (and we do have more in the Agda formalization).
+  and $\elir$ by simultaneous IR:
 \begin{alignat*}{5}
@@
 \end{mydefinition}
+We use the prime accents ($'$) to disambiguate IR codes from type formers in TTGU or the metatheory.
+For basic type formers, we only include codes for function types, the empty type, and $\Bool$.
+Other type formers are straightforward to add (and we do have more in the Agda formalization).
 
@@
 \begin{alignat*}{3}
-  & \ult\,: (i : \Lvl)\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
+  & \ult\,: (i : \Lvl)\{p : \Acc\,(\blank\!<\!\blank)\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
   & \ult\,i\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,j\,\{f\,j\,p\})}
@@
 Assuming $p : i < j$, we have the computation rule $\ult\,j\,p = \U\,j$. Proof: we
-may assume that any witness for $\Acc\,<\,i$ is of the form $\acc\,f$ for
+may assume that any witness for $\Acc\,(\blank\!<\!\blank)\,i$ is of the form $\acc\,f$ for
 some $f$. Then the equation becomes $\uir_{j\,(\ult\,j\,\{f\,j\,p\})} =
@@
 \begin{mydefinition}[Semantic $\Lift$]
-We define by induction on $\uir$ a function with type $(p : i < j)(A : \U\,i)
+We define by induction on $\uir$ a function with type $(p : i < j)\to(A : \U\,i)
 \to (A' : \U\,j)\times(\El\,A' = \El\,A)$. However, for the sake of clarity, we
@@
   &\Lift\,      &&: (p : i < j) \to \U\,i \to \U\,j\\
-  &\msf{ElLift} &&: (p : i < j)(A : \U\,i) \to \El\,(\Lift\,A) = \El\,A
+  &\msf{ElLift} &&: (p : i < j)\to(A : \U\,i) \to \El\,(\Lift\,A) = \El\,A
 \end{alignat*}
@@
 \end{alignat*}
-Above, the $\Pi'$ definition is well-typed by $\msf{ElLift}\,p\,A$. For the
-proof of $\msf{ElLift}$, the only interesting case is $\U'$. Here, we need
-show $\ult\,j\,k\,(p \circ q) = \ult\,i\,k\,q$, but by Lemma
-\ref{lem:ucomp} both sides are $\U\,k$. \qed
+Above, the $\Pi'$ definition is well-typed by $\msf{ElLift}\,p\,A$. For the proof of $\msf{ElLift}$,
+the only interesting case is $\U'$. Here, we need to show $\ult\,j\,k\,(p \circ q) = \ult\,i\,k\,q$,
+but by Lemma \ref{lem:ucomp} both sides are $\U\,k$.
 \end{mydefinition}
@@
 \begin{mydefinition}[Base category]
-This is simply the category of sets and functions in $\Set0$, i.e.\ $\bm{\Con} = \Set0$,
-$\bm{\Sub}\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
+The base category is simply the category of sets and functions in $\Set0$, i.e.\ $\bm{\Con} =
+\Set0$, $\bm{\Sub}\,\Gamma\,\Delta = \Gamma \to \Delta$, and the terminal object is $\top$.
 \end{mydefinition}
@@
 \end{alignat*}
-Type and term substitution are given by composition with some $\sigma : \Gamma
-\to \Delta$ function. Comprehension structure is given by $\Gamma \bm{\ext} A =
+Type and term substitution are given by composition with some function $\sigma : \Gamma
+\to \Delta$. Comprehension structure is given by $\Gamma \bm{\ext} A =
 (\gamma : \Gamma) \times \El\,(A\,\gamma)$. Type lifting along $p : i < j$ is
@@
 \begin{alignat*}{5}
-  & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\Gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
+  & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
   & \hspace{2em}\bm{\bot}_i = \lambda\,\gamma.\,\bot'
@@
   equality, and term lifting is the identity function.
-\item We define $\bm{\U}\,i\,j\,p = \lambda\,\gamma.\,\U'\,(p\,\gamma)$. Again,
+\item We define $\bm{\U}\,i\,j\,p = \lambda\,\gamma.\,\U'\,(i\,\gamma)\,(p\,\gamma)$. Again,
   we have $\bm{\Tm}\,\Gamma\,(\U\,i\,j\,p) = \bm{\Ty}\,\Gamma\,i$ by Lemma \ref{lem:ucomp},
@@
 Sterling \cite{sterling2019algebraic} gave an algebraic specification much like
-ours for a type theory with countable cumulative universes, and prove canonicity
+ours for a type theory with countable cumulative universes, and proved canonicity
 for it.

### 2021-04-08 efop_prez/prez.tex
translate efop prez to Hungarian

@@
 
-\title{Universes In Type Theory}
-\date{April 8, 2021, EFOP Workshop}
-\author{\normalsize{\vspace{-1em}\textbf{András Kovács}\footnote{The author was supported by the European Union, co-financed by the
-European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}}
+\title{Univerzumok Típuselméletben}
+\date{2021 április 8., EFOP Workshop}
+\author{\normalsize{\vspace{-1em}\textbf{Kovács András}\footnote{A szerzőt az Európai Unió támogatta, közös finanszírozásban az Európai Szociális Alappal (EFOP-3.6.3-VEKOP-16-2017-00002).\vspace{0.5em}}}}
 %% \institute{Eötvös Loránd University, Budapest}
@@
 
-\begin{frame}{Russell's paradox, or size issues}
+\begin{frame}{Russell paradoxon és méret-problémák}
 
-There is no set $S$ in a consistent set theory such that $x \in S$ iff $x \notin x$.
+Nincs olyan $S$ halmaz egy konzisztens halmazelméletben, hogy $x \in S$ pontosan akkor, ha $x \notin
+x$.
 \vspace{1em}
 
-Alternate formulations:
+Alternatív fogalmazás:
 \begin{itemize}
-  \item Barber's paradox
-  \item No set of all sets
+  \item Borbély paradoxon
+  \item Összes halmaz halmaza
 \end{itemize}
@@
 
-Related:
+Kapcsolódó:
 \begin{itemize}
-  \item Real numbers are uncountable
-  \item Gödel's first incompleteness theorem
-  \item Undecidability of halting
+  \item Valós számok nem megszámlálhatóak
+  \item Gödel első nem-teljességi tétele
+  \item Megállási probléma
 \end{itemize}
@@
 
-{\footnotesize (All instances of \href{https://arxiv.org/pdf/math/0305282.pdf}{Lawvere's fixed point theorem)}}
+{\footnotesize (Általánosításuk: \href{https://arxiv.org/pdf/math/0305282.pdf}{Lawvere fixpont tétele)}}
 \end{frame}
@@
 
-\begin{frame}{Solution with universes}
+\begin{frame}{Megoldás univerzumokkal}
 
-Solving Russell's paradox: set comprehension can only define a \textbf{subset} of a set.
+Csak \textbf{létező halmaz részhalmazát} definiálhatjuk logikai formulával.
 \vspace{1em}
 
-But sometimes we still want to quantify over \textbf{all sets} of
-Zermelo-Fraenkel set theory.
+De olykor szeretnénk kvantifikálni egy halmazelmélet \textbf{minden halmaza}
+fölött.
 \vspace{1em}
 
-\textbf{Universe:} a set which includes every ZF set.
+\textbf{Univerzum:} halmaz, ami tartalmazza egy halmazelmélet (pl. ZF) összes halmazát.
 \vspace{1em}
 
-Gödel: there is no universe inside ZF.
+Gödel, Russell: ZF-ben nincs univerzum, ami az összes ZF halmazt tartalmazza.
 \vspace{1em}
 
-We can assume more universes if we want to quantify over more sets.
+Új univerzumokat axiómaként kell felvenni.
 \end{frame}
 
-\begin{frame}[fragile]{Universes in type theory}
+\begin{frame}[fragile]{Típuselméletben}
 
-Type theories:
+Típuselméletek:
 \begin{itemize}
-  \item Alternative to set theory as mathematical foundation.
-  \item Used in most general-purpose theorem provers.
-  \item Can be used for proving and programming at the same time.
+  \item A halmazelméletek alternatívái.
+  \item Tételbizonyító-rendszerek gyakori alapja.
+  \item Bizonyításra és programozásra egyszerre alkalmas.
 \end{itemize}
 
-Universes come up in type theories as well.
+Formális vs. informális univerzum kezelés.
 \vspace{1em}
 
-Every theorem prover has different universe-related features.
+Létező bizonyító-rendszerekben: univerzum megvalósítások sokasága kisebb-nagyobb eltérésekkel.
 \vspace{1em}
 
-We want to give a general semantics covering many different features.
+Szeretnénk ezeket lefedni egy általános szemantikával.
 
@@
 
-\begin{frame}[fragile]{Example}
+\begin{frame}[fragile]{Példa}
 
@@
 
-Identity function at all small sets:
+Identitás függvény minden ``kis'' halmazon:
 \begin{verbatim}
@@
 
-We have $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$ for all $i$.
+Minden $i$-re $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$.
 \vspace{1em}
 
-Assuming $\mathsf{Set_i} : \mathsf{Set_{i + 1}}$ implies contradiction (by Russell-like argument).
+Ellentmondás lenne: $\mathsf{Set_i} : \mathsf{Set_{i}}$.
 \vspace{1em}
@@
 
-\begin{frame}[fragile]{Extensions}
+\begin{frame}[fragile]{Kiterjesztések}
 
-The basic system can be tedious:
+Az alap rendszer vesződséges tud lenni:
 \begin{verbatim}
@@
 
-In Coq/Agda/Lean, various extra features are used.
-Example: universe polymorphism in Agda:
+Coq/Agda/Lean-ben extra feature-ök vannak.
+Pl. univerzum polimorfizmus Agda-ban:
 \begin{verbatim}
@@
 
-\begin{frame}[fragile]{Design choices and variations}
+\begin{frame}[fragile]{Variációk}
 
-\textbf{How many universes?} Agda/Coq: countably many.
+\textbf{Mennyi univerzum?} Agda/Coq: megszámlálható.
 \vspace{1em}
 
-\textbf{Are universes totally ordered?} Agda/Coq: yes.
+\textbf{Univerzumok teljesen rendzettek?} Agda/Coq: igen.
 \vspace{1em}
 
-\textbf{What kind of level polymorphism?}. Coq: bounded polymorphism. Agda: no
-bounds allowed. Bounded example:
+\textbf{Milyen univerzum polimorfizmus?}. Coq: korlátos polimorfizmus. Agda: nincsenek
+korlátok. Példa korláttal:
 \begin{verbatim}
@@
 
-\textbf{What kind of operations are available on levels?} Agda is more liberal than Coq. Example:
+\textbf{Milyen műveletek vannak univerzum-szinteken?} Az Agda liberálisabb a Coq-nál:
 \begin{verbatim}
@@
 
-\textbf{Are universes cumulative} Agda: no. Coq: yes. Cumulativity: whenever \texttt{A : Set i}, we also have \texttt{A : Set (i + 1)}.
+\textbf{Kumulativitás.} Agda: nincs. Coq: van. Kumulativitás: ha \texttt{A : Set i}, akkor \texttt{A
+  : Set (i + 1)}.
 
@@
 
-\begin{frame}{Research goals}
+\begin{frame}{Kutatási célok}
 
-We want to know that each point in the design space makes sense.
+Szeretnénk belátni, hogy a különféle variációk mind jól viselkednek.
 \vspace{1em}
 
-Making sense:
+Jól viselkedés:
 \begin{itemize}
-  \item Logical consistency.
-  \item Is the type theory a proper programming language? Programs should compute to values
-        and not get randomly stuck.
-  \item Is proof checking decidable? (Not covered in current work).
+  \item Logikai konzisztencia.
+  \item Igaz, hogy a típuselmélet egy programozási nyelv? Azaz: programok le kell, hogy
+        fussanak elakadás nélkül. (Nincs jelenlegi munkában)
+  \item Eldönthető a bizonyítások validálása? (Nincs jelenlegi munkában)
 \end{itemize}
 
-Approach: use generic framework to cover as many features/variations as
-possible. Prove that everything in the framework is sensible.
+Egy általános rendszert használunk, ahol sok feature modellezhető, és
+egyszerre bizonyítjuk az összes konzisztenciáját.
 
@@
 
-\begin{frame}[fragile]{Features covered by the framework}
+\begin{frame}[fragile]{Amit a rendszerünk lefed}
 
-Universe levels may come from \textbf{any well-ordered set}, even transfinite:
+Univerzum-szintek \textbf{bármilyen jól-rendezett halmazt} alkothatnak.
 \begin{verbatim}
@@
 \end{verbatim}
-\textbf{Quantification over levels}, \textbf{arbitrary computation on levels}.
+\textbf{Szintek fölötti kvantifikáció}, \textbf{tetszőleges számítások szinteken}.
 \begin{verbatim}
@@
 \end{verbatim}
-\textbf{Universes are cumulative}
+\textbf{Kumulativitás}
 \begin{verbatim}
@@
 
-\begin{frame}[fragile]{Implementation}
+\begin{frame}[fragile]{Megvalósítás}
 
-We define a family of type theories with general universe features.
+Visszavezetjük az összes feature-t egy már ismert feature-re,
+az \textbf{induktív-rekurzív} definíciókra.
 \vspace{1em}
 
-We prove all of them consistent by reducing all fancy features to
-a single previously known feature, called \textbf{induction-recursion}.
-\vspace{1em}
-
-Induction-recursion is a type-theoretic analogue of assuming \textbf{Mahlo cardinals} in set theory.
+Az indukció-rekurzió egy típuselméleti változata a halmazelméleti \textbf{Mahlo kardinálisoknak}.
 
@@
 
-%% \begin{enumerate}
-%%   \item Universe levels may come from any well-ordered set.
-%%   \item Polymorphism with bounds is allowed.
-%%   \item Universes are cumulative.
-%%   \item Levels can be manipulated arbitrarily as program data (WIP).
-%% \end{enumerate}
-
-%% Things we need to assume in order to prove the framework sensible:
-%% \begin{enumerate}
-%%   \item \textbf{Two} universes.
-%%   \item Types of certain infinitely branching trees (W-types).
-%% \end{enumerate}
-
-%% Potential further applications (in future work):
-%% \begin{itemize}
-%%   \item Information flow type systems (``secure'' and ``public'' levels).
-%%   \item Staged compilation (``runtime'' and ``compile time'' levels).
-%% \end{itemize}
-
-%% \end{frame}
-
-\begin{frame}{Publication}
+\begin{frame}{Publikáció}
 
@@
 
-Submission under review for FSCD 2021.
+Elbírálás alatt FSCD 2021 konferenciára.
 
@@
 
-\begin{center} {\large Thank you!} \end{center}
+\begin{center} {\large Köszönöm a figyelmet!} \end{center}
 

### 2021-04-19 arxivsub/preprint.tex
typo
update arxiv sub with fixes

@@
 \nolinenumbers %uncomment to disable line numbering
-
 \hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
@@
 
+\usepackage{xcolor}
+\usepackage{mathpartir}
+\usepackage{todonotes}
+\presetkeys{todonotes}{inline}{}
 \usepackage{scalerel}
+\usepackage{amssymb}
 \usepackage{bm}
-\usepackage{hyperref}
+
+%% \theoremstyle{definition}
+%% \newtheorem{definition}{Definition}
+%% \newtheorem{example}{Example}
+
+%% \theoremstyle{theorem}
+%% \newtheorem{theorem}{Theorem}
+%% \newtheorem{lemma}{Lemma}
 
@@
 
-We provide an Agda formalization of the contents of the paper, at
-\url{https://github.com/AndrasKovacs/universes/tree/master/agda}. The
-formalization is not complete, as we skip proofs involving an excessive number
-of equality coercions (which are more suited to informal reasoning, using
-equality reflection), and instead focus on the key points.
+We provide an Agda formalization of the contents of the paper. The formalization
+is not complete, as we skip proofs involving an excessive number of equality
+coercions (which are more suited to informal reasoning, using equality
+reflection), and instead focus on the key points.
 

### 2021-06-05 arxivsub/preprint.tex
fix

@@
 \end{mylemma}
-\begin{proof} From the
-preservation of context extension by $F$, we know that $F$ is invertible on the zero De Bruijn
-index, so for $\q : \Tm\,(\Gamma \ext F\,A)\,(F (A[\p]))$, we have $F^{-1}\q : \Tm\,(\Gamma \ext
-F\,A)\,(A[\p])$, such that $F (F^{-1}\,\q) = \q$. If we have any $t : \Tm\,\Gamma\,(F\,A)$, we also
-have $(F^{-1}\q)[\id,\,t] : \Tm\,\Gamma\,A$, and this yields the inverse action $F^{-1}$ which acts
-on all terms.
+
+\begin{proof}
+From the $\ext$-preservation isomorphism and the defining isomorphisms of
+comprehension, we get $\q' : \Tm\,(\Gamma \ext F\,A)\,(A[\p])$ such that $F\,\q'
+= \q$ and $\q'[\p,\,F\,\q] = \q$. Now, for $t : \Tm\,\Gamma\,(F\,A)$, we
+define $F^{-1}\,t$ as $\q'[\id,\,t] : \Tm\,\Gamma\,A$. We get the following:
+\begin{alignat*}{3}
+  & F(F^{-1}\,t) = F(\q'[\id,\,t]) = (F\,\q')[\id,\,t] = \q'[\id,\,t] = t \\
+  & F^{-1}(F\,t) = \q'[\id,\,F\,t] = \q'[\id,\,(F\,\q)[\id,\,\,t]] = \q'[\p,\,F\,\q][\id,\,t]
+    = \q[\id,\,t] = t
+\end{alignat*}
+More concisely, $F$ is invertible on the generic term $\q$, which implies invertibility
+on any term.
 \end{proof}

### 2021-07-12 paper.tex
new sub minor fixes

@@
 
-We provide an Agda formalization of the contents of the paper. The formalization
-is not complete, as we skip proofs involving an excessive number of equality
-coercions (which are more suited to informal reasoning, using equality
-reflection), and instead focus on the key points.
+We provide an Agda formalization of the contents of the paper at
+\url{https://github.com/AndrasKovacs/universes/tree/master/agda}. The
+formalization is not complete, as we skip proofs involving an excessive number
+of equality coercions (which are more suited to informal reasoning, using
+equality reflection), and instead focus on the key points.
 
@@
 use the cwf presentation for its immediately algebraic character and closeness
-to conventional explicit substitution syntax.
+to conventional explicit substitutions. We consider the syntax of a type theory
+to be its initial model.
 
@@
 
-Second, could we use weaker type-theoretic features, like super universes
+Second, could we use simpler type-theoretic features, like super universes
 \cite{Palmgren98onuniverses} or Mahlo universes \cite{setzer00mahlo}? These are

### 2021-10-29 paper.tex
fixes

@@
 \newcommand{\false}{\mathsf{false}}
-\newcommand{\up}{\uparrow}
-\newcommand{\down}{\downarrow}
+\newcommand{\up}{{\uparrow}}
+\newcommand{\down}{{\downarrow}}
 \newcommand{\Lift}{\mathsf{Lift}}
@@
 that it is natural with respect to substitution, i.e.\ all type and term formers
-have a corresponding substitution rule.
+have a corresponding substitution rule. This convention could be made precise by
+working in a framework for higher-order abstract syntax, where all specified
+structure is automatically stable under substitution
+\cite{cubicalnorm,uemura,bocquet2021relative}. While this can be effective at
+reducing formal clutter, this paper only presents models which are technically
+straightforward, so we choose not to use higher-order signatures, in order to
+make the presentation more direct.
 
@@
 \begin{alignat*}{3}
-  & F(F^{-1}\,t) = F(\q'[\id,\,t]) = (F\,\q')[\id,\,t] = \q'[\id,\,t] = t \\
+  & F(F^{-1}\,t) = F(\q'[\id,\,t]) = (F\,\q')[\id,\,t] = \q[\id,\,t] = t \\
   & F^{-1}(F\,t) = \q'[\id,\,F\,t] = \q'[\id,\,(F\,\q)[\id,\,\,t]] = \q'[\p,\,F\,\q][\id,\,t]
@@
 
-Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a morphism between them. This
-corresponds to a basic version of \emph{two-level type theory} \cite{twolevel}. This theory can be
-interpreted as having ($\Ty_1$, $\Tm_1$) as a metaprogramming layer which can generate object-level
-programs in the ($\Ty_0$, $\Tm_0$) layer. Lifted types correspond to types of object-level
-expressions; for example, $\Bool_0 : \Ty_0\,\Gamma$ is the object-level type of Booleans, while
-$\Lift\,\Bool_0$ is the meta-level type of $\Bool_0$-expressions, and $\Bool_1 : \Ty_1\,\Gamma$ is
-the type of meta-level Booleans. It is possible to compute a $\Bool_0$ from a $\Bool_1$. Given $b :
-\Tm_1\,\Gamma\,\Bool_1$, we can construct
-$\down(\mathsf{if}\,b\,\mathsf{then}\up\!\true_0\,\mathsf{else}\up\!\false_0) :
-\Tm_0\,\Gamma\,\Bool_0$.
-
-But there is no way to compute a $\Bool_1$ from a $\Bool_0$: we can try to lift
-the input, but there is no elimination rule for $\Lift\,\Bool_0$ in
-$\Ty_1$. Conceptually, terms of $\Lift\,\Bool_0$ are \emph{expressions}, so they
-are not necessarily $\true$ or $\false$, they can also be variables or neutral
-expressions, so the usual Boolean elimination is not justified for them. In
-general, the setup ensures that we can map from positive types in $\Ty_1$
-to their counterparts in $\Ty_0$, but not the other way around, while negative
-types are preserved by $\Lift$ up to type isomorphism.
-
-The rules of two-level type theory appear to model a form of generative two-stage compilation with
-dependent types. Comparing this system to e.g.\ BER MetaOCaml \cite{kiselyov14metaocaml}, we can
-relate $\Lift$ to $\mathsf{code}$, $\up$ to the quasi-quotation operation $<\!\blank\!>$, and
-$\down$ to escape $\sim\!\blank$.
-
-While the staging aspect could be subject of future research, we currently focus
-on ``sizing'' hierarchies instead of staging hierarchies. This means that we
-want to eliminate from any family to any other family which is connected by
-a morphism.
+Assume family structures ($\Ty_0$, $\Tm_0$) and ($\Ty_1$, $\Tm_1$) and a
+morphism between them. This corresponds to a basic version of \emph{two-level
+type theory} \cite{twolevel}. This theory has an interpretation in presheaves
+over the category of contexts of some chosen model of a type theory, where
+($\Ty_0$, $\Tm_0$) is modeled using structure in the chosen model, and ($\Ty_1$,
+$\Tm_1$) is modeled using presheaf constructions. More illustratively, this
+means interpreting ($\Ty_1$, $\Tm_1$) as a metaprogramming layer which can
+generate object-level constructions in the ($\Ty_0$, $\Tm_0$) layer. Lifted
+types correspond to types of object-level terms; for example, $\Bool_0 :
+\Ty_0\,\Gamma$ is the object-level type of Booleans, while $\Lift\,\Bool_0$ is
+the meta-level type of $\Bool_0$-terms, and $\Bool_1 : \Ty_1\,\Gamma$ is the
+type of meta-level Booleans. It is possible to compute a $\Bool_0$ from a
+$\Bool_1$. Given $b : \Tm_1\,\Gamma\,\Bool_1$, we can construct
+$\down(\mathsf{if}\,b\,\mathsf{then}\,\up\true_0\,\mathsf{else}\,\up\false_0) :
+\Tm_0\,\Gamma\,\Bool_0$. But there is no way to compute a $\Bool_1$ from a
+$\Bool_0$: we can try to lift the input, but there is no elimination rule for
+$\Lift\,\Bool_0$ in $\Ty_1$.
+
+Hence, plain family morphisms can model a metaprogramming hierarchy, but
+currently we are aiming for ``sizing'' hierarchies instead. This means that we
+want to eliminate from any family to any other family which is connected by a
+morphism.
 
@@
 \begin{alignat*}{3}
-  & \Lift\,(\Pi\,(x : A) B)   && =\,\,\,\,&& \Pi\,(x : \Lift\,A)(\Lift\, (B[x \mapsto\,\down\!x]))\\
-  & \up(\lambda\,(x : A).\,t) && =&& \lambda\,(x : \Lift\,A).\up(t[x\mapsto\down\!x])\\
+  & \Lift\,(\Pi\,(x : A) B)   && =\,\,\,\,&& \Pi\,(x : \Lift\,A)(\Lift\, (B[x \mapsto\,\down x]))\\
+  & \up(\lambda\,(x : A).\,t) && =&& \lambda\,(x : \Lift\,A).\,\up(t[x\mapsto\down x])\\
   & \Lift\,\Bool_0            && =&& \Bool_1\\
@@
 eliminate from $\Bool_0$ to $\Bool_1$. If we have some $b :
-\Tm_0\,\Gamma\,\Bool_0$, we also have $\up\!b :
-\Tm_1\,\Gamma\,(\Lift\,\Bool_0)$, hence $\up\!b :
+\Tm_0\,\Gamma\,\Bool_0$, we also have $\up b :
+\Tm_1\,\Gamma\,(\Lift\,\Bool_0)$, hence $\up b :
 \Tm_1\,\Gamma\,\Bool_1$. Then, we can use $\Bool_1$ elimination, as in
-$\mathsf{if}\,\up\!b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
+$\mathsf{if}\,\up b\,\,\mathsf{then}\,\true_1\,\mathsf{else}\,\false_1 :
 \Tm_1\,\Gamma\,\Bool_1$. The $\up$ computation ensures that the eliminator
 computes appropriately on canonical terms: if $b$ is $\true_0$, we get
-$\up\!\true_0 = \true_1$ as the if-then-else scrutinee.
+$\up \true_0 = \true_1$ as the if-then-else scrutinee.
 
@@
   & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
-  & \up\!t &&= t                                    \label{eq:cumlift}
+  & \up t &&= t                                    \label{eq:cumlift}
 \end{alignat}
@@
 hence $\true_0 : \Tm_0\,\Gamma\,\Bool_1$. Moreover, $\true_0$ is definitionally
-equal to $\true_1$, since $\true_0 =\,\up\!\true_0 = \true_1$. Thus, using
+equal to $\true_1$, since $\true_0 =\,\up \true_0 = \true_1$. Thus, using
 simply $\true$ is fine whenever the family is clear from context.
@@
 not sufficient to specify a strict inclusion just by equations \ref{eq:cumcon}
-and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}, we need $\up$
+and \ref{eq:cumtm} in Definition \ref{def:strict_inclusion}. We need $\up$
 together with equation \ref{eq:cumlift} to identify term formers in different
-families. The other direction $\down\!t = t$ is immediately derivable.
+families. The other direction $\down t = t$ is immediately derivable.
 
@@
 between them. In this subsection we describe the indexing structures for such
-family diagrams. First, some preliminary definitions.
-
-\begin{mydefinition} The \emph{accessibility predicate} on relations is defined by the following
+family diagrams. First, we specify a notion of well-foundedness, which will be
+used to preclude size paradoxes in universe hierarchies.
+
+\begin{mydefinition}
+The \emph{accessibility predicate} on relations is defined by the following
 inductive rules:
@@
 \end{mydefinition}
-
-\noindent An inhabitant of $\Acc\,R\,a$ proves that starting from $a : A$, all
-descending $R$-chains must be finite. This is ensured by the universal property
-of the inductive definition, as all inductive types must be well-founded.
+\noindent See \cite{aczel1977introduction} and \cite[Section~10.3]{hottbook} for
+further exposition. An inhabitant of $\Acc\,R\,a$ proves that starting from $a :
+A$, all descending $R$-chains must be finite. This is ensured by the universal
+property of the inductive definition.
 
 \begin{mylemma}\label{lem:accprop}
-Accessibility is a mere proposition, i.e.\ all inhabitants of $\Acc\,R\,a$ are
-equal \cite[Lemma 10.3.4]{hottbook}.
+All inhabitants of $\Acc\,R\,a$ are equal \cite[Lemma 10.3.4]{hottbook}. In
+other words, accessibility is proof-irrelevant.
 \end{mylemma}
@@
 can internally prove them isomorphic using function types and identity types. That said, every
-construction in this paper would still work with inverse categories as level structures.
+construction in this paper would still work with direct categories as level structures.
 
@@
 former depends on levels in an interesting way. For example, $\Bool_i$ has the
-same canonical inhabitants as $\Bool_j$, for any $i$ and $j$. Universes
-introduce dependency on levels, by serving as classifiers for smaller families
-internally to larger families.
+same inhabitants as $\Bool_j$, for any $i$ and $j$. Universes introduce
+dependency on levels, by serving as classifiers for smaller families internally
+to larger families.
 
@@
 \begin{mydefinition}
-A family diagram has \emph{Coquand universes} if it has universe formation and
+A family diagram has \emph{Coquand universes} \cite{coquandnorm} if it has universe formation and
 additionally supports $\El : \Tm_j\,\Gamma\,(\U\,i\,j\,p) \to \Ty_i\,\Gamma$, and
@@
 In this section we give a model for a type theory with generalized
-universes. Let's make the notion of model concrete first.
+universes. Let us make the notion of model concrete first.
 
@@
 
-The task is to interpret the $\Lvl$-many universes of TTGU using a metatheoretic
-feature which is more compact and already well-studied. For this, we need to
-define a $\Lvl$-indexed type of type codes. Since $\Lvl$ and $\!\blank<\!\blank$
-can be arbitrary, we effectively need to define transfinite hierarchies of
-codes. We use an inductive-recursive (IR) \cite{dybjer99finite} definition, for
-the following reasons.
-
-First, IR is already supported in the proof assistant Agda, and it is very
-useful to be able to sketch out ideas in a machine-checked setting. It would be
-much harder to do the same when developing semantics in set theory.
-
-Second, could we use simpler type-theoretic features, like super universes
-\cite{Palmgren98onuniverses} or Mahlo universes \cite{setzer00mahlo}? These are
-sufficient to model transfinite hierarchies. However, it is not clear how to
-additionally support \emph{recursive sub-universes}, which in our work corresponds to
-the type former preservation property of $\Lift$. Thus, we give a custom IR
-definition which corresponds more directly to TTGU structure.
-
-\begin{mydefinition}[IR codes]
-  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$
-  and $\elir$ by simultaneous IR:
+The task is to interpret the $\Lvl$-many universes of TTGU using an assumed
+metatheoretic feature. For this, we need to define a $\Lvl$-indexed type of type
+codes. Since $\Lvl$ and $\!\blank<\!\blank$ can be arbitrary, we effectively
+need to define transfinite hierarchies of codes. We use an inductive-recursive
+\cite{dybjer99finite} definition for the following reasons.
+
+First, induction-recursion is already supported in the Agda proof assistant, and
+it is very useful to be able to sketch out ideas in a machine-checked
+setting. It would be much harder to do the same when developing semantics in set
+theory.
+
+Second, could we use type-theoretic features with simpler specifications than
+induction-recursion, such as super universes \cite{Palmgren98onuniverses} or
+Mahlo universes \cite{setzer00mahlo}? These are sufficient to model transfinite
+hierarchies. However, using these it is not clear how to additionally support
+the strict type former preservation property of $\Lift$\footnote{Palmgren calls
+this property as having \emph{recursive sub-universes}
+\cite{Palmgren98onuniverses}.}.
+
+Therefore, we give a custom definition using induction-recursion, which
+corresponds more directly to TTGU structure. Our definition is essentially the
+same as McBride's redundancy-free hierarchy in
+\cite[Section~6.3.1]{mcbride2015datatypes}, but we generalize levels from
+natural numbers to arbitrary level structures.
+
+\begin{mydefinition}[Codes for the universe]
+Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$ and
+$\elir$ by induction-recursion:
 \begin{alignat*}{5}
@@
 \end{mydefinition}
-We use the prime accents ($'$) to disambiguate IR codes from type formers in TTGU or the metatheory.
-For basic type formers, we only include codes for function types, the empty type, and $\Bool$.
-Other type formers are straightforward to add (and we do have more in the Agda formalization).
+We use the prime accents ($'$) to disambiguate inductive-recursive codes from
+type formers in TTGU or the metatheory. For basic type formers, we only include
+codes for function types, the empty type, and $\Bool$. Other type formers are
+straightforward to add (and we do have more in the Agda formalization).
 
@@
 
-\subsection{IR Model of TTGU}
+\subsection{Inductive-Recursive Model of TTGU}
 
@@
 \item
-  A ``bootstrapping'' assumption on levels. This can be any collection of levels
-  and morphisms. We pick the assumption that $l_0, l_1 : \Lvl\,\Gamma$ exist
-  together with $l_{01} : l_0 < l_1$. We need to assume at least some levels,
-  because otherwise the TTDL syntax is trivial; it is not even possible to
-  write down a non-empty context or a closed type former. $l_0 < l_1$ allows large
-  eliminations on type formers so it provides a fair amount of expressive power.
+  A ``bootstrapping'' assumption on levels. This can be any non-empty collection
+  of levels and morphisms. It will be used shortly in Section
+  \ref{sec:level_reflection}, where we specify first-class levels using the syntax
+  (i.e.\ the initial model) of TTDL. Without bootstrapping, the syntax is
+  trivial and has no closed types. Of course, models of TTDL in general make
+  sense without the bootstrapping assumption.
+
+  We pick the assumption that $l_0, l_1 : \Lvl\,\Gamma$ exist together with
+  $l_{01} : l_0 < l_1$. This allows large eliminations on type formers, so it
+  provides a fair amount of power for specifying internal levels.
 \item
@@
   \end{alignat*}
-  We have type and term substitution, which depends on level substitution. For instance, we have
+  We have type and term substitution, which depends on level substitution. For instance, we have:
   \[
@@
 
-  Above we mention basic type formers, although we haven't yet specified
-  those. The way this should be understood, is that any basic type former introduced
-  from now on should come equipped with preservation equations for lifting. This is
-  similar to how we mandate that any introduced type former must be natural with respect
-  to substitution.
+  Above we mention basic type formers, although we have not yet specified
+  those. The way this should be understood, is that any basic type former
+  introduced from now on should come equipped with preservation equations for
+  lifting. This is similar to how we mandate that any introduced type former
+  must be natural with respect to substitution.
 \item A universe structure
@@
     &\U  &&: \{\Gamma : \Con\}(i\,j : \Lvl\,\Gamma) \to i < j \to \Ty\,\Gamma\,j\hspace{2em}
-    &\El &&: \Tm\,\Gamma\,(\U\,i\,j\,p) \to \Ty\,\Gamma\,i
-  \end{alignat*}
-  Such that $\Lift\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)$, $\El$ has
+    &\El\,\,&&: \Tm\,\Gamma\,(\U\,i\,j\,p) \to \Ty\,\Gamma\,i
+  \end{alignat*}
+  such that $\Lift\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)$, $\El$ has
   inverse $\Code$, $\Tm\,\Gamma\,(\U\,i\,j\,p) = \Ty\,\Gamma\,i$ and $\El\,t =
@@
 
-\begin{mydefinition}[IR model of TTDL]\label{def:ttdlmodel}
- Assume an external $\Lvl$ structure such that it supports $l_0, l_1 : \Lvl$ and
- $l_{01} : l_0 < l_1$ (the bootstrapping assumption). We again use the IR
- universe constructions from Section \ref{sec:inductive_recursive_codes},
- instantiated to the assumed $\Lvl$ structure. We describe components of the
- model in order. Again, we write components of the model in \textbf{bold} font.
+\begin{mydefinition}[Inductive-recursive model of TTDL]\label{def:ttdlmodel}
+ Assume an external $\Lvl$ structure that supports $l_0, l_1 : \Lvl$ and $l_{01}
+ : l_0 < l_1$ (the bootstrapping assumption). We again use the universe
+ constructions from Section \ref{sec:inductive_recursive_codes}, instantiated to
+ the assumed $\Lvl$ structure. We describe components of the model in
+ order. Again, we write components of the model in \textbf{bold} font.
 \begin{enumerate}
@@
 To summarize, the only interesting change compared to the TTGU model is that
-levels and level morphisms gain potential dependency on contexts. However, in the
-IR model, this is simply the addition of an extra semantic function parameter.
+levels and level morphisms gain potential dependency on contexts. However, in
+the inductive-recursive model this is simply the addition of an extra semantic
+function parameter.
 
@@
 for a type theory with first-class levels (TTFL). We can easily get a TTFL with
-an IR model in the following way. First, pick an external $\Lvl$ structure which
-a) satisfies the bootstrapping assumption b) has sets of levels and morphisms
-which can be represented with syntactic TTDL types.
+an inductive-recursive model in the following way. First, we pick an external
+$\Lvl$ structure which a) satisfies the bootstrapping assumption b) has sets of
+levels and morphisms which can be represented with syntactic TTDL types.
 
@@
 
-Note that $\Pi$ is contravariant in the domain. This is easily supported
-with our IR semantics, unlike in the set-theoretic model of cumulativity for Coq
-\cite{timany18cumulative}, where function domains are invariant.
+Note that $\Pi$ is contravariant in the domain. This is easily supported with
+our inductive-recursive semantics, unlike in the set-theoretic model of
+cumulativity for Coq \cite{timany18cumulative}, where function domains are
+invariant.
 
@@
 case. For example, Agda 2.6.1 has $\omega + 1$ levels (externally), where
-$\Set{\omega}$ is the topmost universe, but $\Set{\omega}$ has no type.
+$\Set{\omega}$ is the topmost universe, but $\Set{\omega}$ is not in any
+universe.
 
@@
   indexedir, positiveir} are particularly flexible and powerful extensions to
-universes.
+universes. McBride gave an inductive-recursion definition of cumulative
+universes that we adapted in this work \cite{mcbride2015datatypes}.
 
@@
 In the current work, we developed a framework for modeling a variety of universe
-features in type theories. At this point, we may ask the question: if IR is
-sufficient to model every feature, why not simply add IR support in a practical
-implementation, and drop the menagerie of universe features?
-
-The answer is that IR provides is a \emph{deep embedding} of universe features,
-which is usually less convenient to use than \emph{native} features. For example,
-both Coq and Agda have powerful automatic solving for filling out implicit
-universe levels. We also do not have to invoke $\El$ or the $\ult$ computation
-rule explicitly, and in Coq we can use implicit syntax for subtyping instead of
-explicit coercions.
+features in type theories. At this point, we may ask the question: if
+induction-recursion is sufficient to model every feature, why not simply support
+it in a practical implementation, and drop the menagerie of universe features?
+
+The answer is that induction-recursion provides a \emph{deep embedding} of
+universe features, which is usually less convenient to use than \emph{native}
+features. For example, both Coq and Agda have powerful automatic solving for
+filling out implicit universe levels. We also do not have to invoke $\El$ or the
+$\ult$ computation rule explicitly, and in Coq we can use implicit syntax for
+subtyping instead of explicit coercions.
 
@@
 usual proof method of gluing (in other words, proof-relevant logical predicates)
-\cite{kaposi2019gluing, sterling2019algebraic} can be adapted to the theories in
+\cite{coquandnorm,kaposi2019gluing,sterling2019algebraic} can be adapted to the theories in
 this paper.
@@
 Lastly, we do not handle impredicative universes. The main reason for this is
-that we do not know the consistency of having IR and impredicative function
-space together in the same universe, and modeling impredicativity seems to
-require this assumption in the metatheory. This could be investigated as well
-in future work.
+that we do not know the consistency of having induction-recursion and
+impredicative function space together in the same universe, and modeling
+impredicativity seems to require this assumption in the metatheory. This could
+be investigated as well in future work.
 
